
\begin{samepage}
{\PaliColumn{1. kāmasuttaṃ (KN 5.39)}}\\
{\setstretch{.7}
\quad\EnglishColumn{Sensual Pleasures}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{772.}}\\

\begin{samepage}
{\PaliColumn{kāmaṃ kāmayamānassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{When a mortal desires sensual pleasure,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tassa ce taṃ samijjhati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{if that succeeds for him,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{addhā pītimano hoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{surely he is elated in mind,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{laddhā macco yadicchati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{having gained what he desires.}\
\par}
\end{samepage}

{\PaliColumn{773.}}\\

\begin{samepage}
{\PaliColumn{tassa ce kāmayānassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But when, full of sensual desire,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{chandajātassa jantuno;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a person is aroused by desire,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{te kāmā parihāyanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{if those sensual pleasures decline,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sallaviddhova ruppati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he is afflicted as if pierced by a dart.}\
\par}
\end{samepage}

{\PaliColumn{774.}}\\

\begin{samepage}
{\PaliColumn{yo kāme parivajjeti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One who avoids sensual pleasures}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sappasseva padā siro;}}\\
{\setstretch{.7}
\quad\EnglishColumn{like a serpent’s head with the foot,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{somaṃ visattikaṃ loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{being mindful, overcomes}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sato samativattati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{this attachment to the world.}\
\par}
\end{samepage}

{\PaliColumn{775.}}\\

\begin{samepage}
{\PaliColumn{khettaṃ vatthuṃ hiraññaṃ vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Fields, land, or bullion,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{gavassaṃ dāsaporisaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{cattle and horses, slaves and servants,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{thiyo bandhū puthu kāme,}}\\
{\setstretch{.7}
\quad\EnglishColumn{women, relatives — when a person greedily}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yo naro anugijjhati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{longs for various sensual pleasures,}\
\par}
\end{samepage}

{\PaliColumn{776.}}\\

\begin{samepage}
{\PaliColumn{abalā naṃ balīyanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{even the weak overpower him;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{maddantenaṃ parissayā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{obstacles crush him.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tato naṃ dukkhamanveti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Then suffering enters him}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāvaṃ bhinnamivodakaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{as water does a broken boat.}\
\par}
\end{samepage}

{\PaliColumn{777.}}\\

\begin{samepage}
{\PaliColumn{tasmā jantu sadā sato,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Therefore, ever mindful, a person}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāmāni parivajjaye;}}\\
{\setstretch{.7}
\quad\EnglishColumn{should avoid sensual pleasures.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{te pahāya tare oghaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having abandoned them, one can cross the flood,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāvaṃ sitvāva pāragūti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{as by bailing out the boat one can reach the far shore.}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{kāmasuttaṃ paṭhamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{2. guhaṭṭhakasuttaṃ (KN 5.40)}}\\
{\setstretch{.7}
\quad\EnglishColumn{The Octad on the Cave}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{778.}}\\

\begin{samepage}
{\PaliColumn{satto guhāyaṃ bahunābhichanno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Stuck in the cave, densely covered over,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tiṭṭhaṃ naro mohanasmiṃ pagāḷho;}}\\
{\setstretch{.7}
\quad\EnglishColumn{dwelling immersed in bewilderment,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dūre vivekā hi tathāvidho so,}}\\
{\setstretch{.7}
\quad\EnglishColumn{a person such as this is far from seclusion,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāmā hi loke na hi suppahāyā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for in the world sensual pleasures are not easily abandoned.}\
\par}
\end{samepage}

{\PaliColumn{779.}}\\

\begin{samepage}
{\PaliColumn{icchānidānā bhavasātabaddhā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Based upon desire, bound to enjoyment of existence,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{te duppamuñcā na hi aññamokkhā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{they let go with difficulty, for there is no release through others.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pacchā pure vāpi apekkhamānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Concerned, too, with the future or the past,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{imeva kāme purimeva jappaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{they hanker for these sensual pleasures or earlier ones.}\
\par}
\end{samepage}

{\PaliColumn{780.}}\\

\begin{samepage}
{\PaliColumn{kāmesu giddhā pasutā pamūḷhā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Those greedy for sensual pleasures, intent on them,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{avadāniyā te visame niviṭṭhā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{are confused, stingy, settled in the unrighteous.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dukkhūpanītā paridevayanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{When they come upon suffering they lament:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kiṃsū bhavissāma ito cutāse.}}\\
{\setstretch{.7}
\quad\EnglishColumn{“What will we be after we pass away here?”}\
\par}
\end{samepage}

{\PaliColumn{781.}}\\

\begin{samepage}
{\PaliColumn{tasmā hi sikkhetha idheva jantu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Therefore a person should train right here:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yaṃ kiñci jaññā visamanti loke;}}\\
{\setstretch{.7}
\quad\EnglishColumn{whatever in the world one might know as unrighteous,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na tassa hetū visamaṃ careyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not on its account act unrighteously,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{appañhidaṃ jīvitamāhu dhīrā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for the wise say this life is short.}\
\par}
\end{samepage}

{\PaliColumn{782.}}\\

\begin{samepage}
{\PaliColumn{passāmi loke pariphandamānaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{I see in the world this population trembling all over,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pajaṃ imaṃ taṇhagataṃ bhavesu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{addicted to craving for states of existence;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hīnā narā maccumukhe lapanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{inferior people prattle in the mouth of death,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{avītataṇhāse bhavābhavesu.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not devoid of craving for various states of existence.}\
\par}
\end{samepage}

{\PaliColumn{783.}}\\

\begin{samepage}
{\PaliColumn{mamāyite passatha phandamāne,}}\\
{\setstretch{.7}
\quad\EnglishColumn{See them trembling over things taken as “mine”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{maccheva appodake khīṇasote;}}\\
{\setstretch{.7}
\quad\EnglishColumn{like fish in a depleted stream with little water.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etampi disvā amamo careyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having seen this too, one should take nothing as “mine,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhavesu āsattimakubbamāno.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not forming attachment to states of existence.}\
\par}
\end{samepage}

{\PaliColumn{784.}}\\

\begin{samepage}
{\PaliColumn{ubhosu antesu vineyya chandaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having removed desire for both ends,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{phassaṃ pariññāya anānugiddho;}}\\
{\setstretch{.7}
\quad\EnglishColumn{having fully understood contact, without greed,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yadattagarahī tadakubbamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{not doing anything for which one might blame oneself,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na lippatī diṭṭhasutesu dhīro.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the wise person is not tainted by things seen or heard.}\
\par}
\end{samepage}

{\PaliColumn{785.}}\\

\begin{samepage}
{\PaliColumn{saññaṃ pariññā vitareyya oghaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having fully understood perception, one can cross the flood.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pariggahesu muni nopalitto;}}\\
{\setstretch{.7}
\quad\EnglishColumn{The muni, untainted by possessions,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{abbūḷhasallo caramappamatto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{with the dart extracted, living heedfully,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāsīsatī lokamimaṃ parañcāti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{does not desire this world or another.}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{guhaṭṭhakasuttaṃ dutiyaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{3. duṭṭhaṭṭhakasuttaṃ (KN 5.41)}}\\
{\setstretch{.7}
\quad\EnglishColumn{The Octad on the Hostile}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{786.}}\\

\begin{samepage}
{\PaliColumn{vadanti ve duṭṭhamanāpi eke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Some speak with hostile minds,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{athopi ve saccamanā vadanti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{while some speak with minds bent on truth.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vādañca jātaṃ muni no upeti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{The muni does not become involved in an arisen dispute;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā munī natthi khilo kuhiñci.}}\\
{\setstretch{.7}
\quad\EnglishColumn{therefore the muni has no barrenness anywhere.}\
\par}
\end{samepage}

{\PaliColumn{787.}}\\

\begin{samepage}
{\PaliColumn{sakañhi diṭṭhiṃ kathamaccayeyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{How could one transcend one’s own view}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{chandānunīto ruciyā niviṭṭho;}}\\
{\setstretch{.7}
\quad\EnglishColumn{if, drawn by desire, one is entrenched in a preference?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sayaṃ samattāni pakubbamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Taking one’s own [view] to be perfect,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yathā hi jāneyya tathā vadeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one would speak as one understands.}\
\par}
\end{samepage}

{\PaliColumn{788.}}\\

\begin{samepage}
{\PaliColumn{yo attano sīlavatāni jantu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{When a person, without being asked, proclaims}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anānupuṭṭhova paresa pāva;}}\\
{\setstretch{.7}
\quad\EnglishColumn{to others his own good behavior and observances,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anariyadhammaṃ kusalā tamāhu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{the skilled say he is one of ignoble nature,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yo ātumānaṃ sayameva pāva.}}\\
{\setstretch{.7}
\quad\EnglishColumn{since on his own accord he proclaims himself.}\
\par}
\end{samepage}

{\PaliColumn{789.}}\\

\begin{samepage}
{\PaliColumn{santo ca bhikkhu abhinibbutatto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But when a peaceful bhikkhu, one inwardly quenched,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{itihanti sīlesu akatthamāno;}}\\
{\setstretch{.7}
\quad\EnglishColumn{does not boast of his good behavior by saying, “I am thus,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tamariyadhammaṃ kusalā vadanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{the skilled say he is one of noble nature}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yassussadā natthi kuhiñci loke.}}\\
{\setstretch{.7}
\quad\EnglishColumn{who has no swellings anywhere in the world.}\
\par}
\end{samepage}

{\PaliColumn{790.}}\\

\begin{samepage}
{\PaliColumn{pakappitā saṅkhatā yassa dhammā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One who has formulated impure teachings,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{purakkhatā santi avīvadātā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{put them together and advanced them,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yadattani passati ānisaṃsaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{becomes attached to the benefit he sees for himself,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{taṃ nissito kuppapaṭicca santiṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{to that peace dependent on the unstable.}\
\par}
\end{samepage}

{\PaliColumn{791.}}\\

\begin{samepage}
{\PaliColumn{diṭṭhīnivesā na hi svātivattā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Attachments to views are not easily overcome;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dhammesu niccheyya samuggahītaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{having decided among teachings, one tightly grasps [a view].}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā naro tesu nivesanesu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Therefore, among those bases of attachment,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nirassatī ādiyatī ca dhammaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{a person rejects and takes up a teaching.}\
\par}
\end{samepage}

{\PaliColumn{792.}}\\

\begin{samepage}
{\PaliColumn{dhonassa hi natthi kuhiñci loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One who is cleansed formulates no view}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pakappitā diṭṭhi bhavābhavesu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{anywhere in the world about various states of existence.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{māyañca mānañca pahāya dhono,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having abandoned hypocrisy and conceit, through what}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa kena gaccheyya anūpayo so.}}\\
{\setstretch{.7}
\quad\EnglishColumn{would the cleansed one go [astray]194 when he is uninvolved?}\
\par}
\end{samepage}

{\PaliColumn{793.}}\\

\begin{samepage}
{\PaliColumn{upayo hi dhammesu upeti vādaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One involved is embroiled in disputes about teachings;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anūpayaṃ kena kathaṃ vadeyya;}}\\
{\setstretch{.7}
\quad\EnglishColumn{but how, about what, could one dispute with one uninvolved?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{attā nirattā na hi tassa atthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Nothing is taken up or rejected by him;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{adhosi so diṭṭhimidheva sabbanti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he has shaken off all views right here.}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{duṭṭhaṭṭhakasuttaṃ tatiyaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{4. suddhaṭṭhakasuttaṃ (KN 5.42)}}\\
{\setstretch{.7}
\quad\EnglishColumn{The Octad on the Pure}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{794.}}\\

\begin{samepage}
{\PaliColumn{passāmi suddhaṃ paramaṃ arogaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“I see the pure, the supreme health;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhena saṃsuddhi narassa hoti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a person achieves full purity through what is seen.”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evābhijānaṃ paramanti ñatvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Directly knowing thus, having known “[It is] supreme,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{suddhānupassīti pacceti ñāṇaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{“I contemplate the pure,” one falls back on knowledge.}\
\par}
\end{samepage}

{\PaliColumn{795.}}\\

\begin{samepage}
{\PaliColumn{diṭṭhena ce suddhi narassa hoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{If a person gains purity by the seen,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñāṇena vā so pajahāti dukkhaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{or if one abandons suffering by knowledge,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aññena so sujjhati sopadhīko,}}\\
{\setstretch{.7}
\quad\EnglishColumn{then one with acquisitions is purified by another;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhī hi naṃ pāva tathā vadānaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the view proclaims him as one who asserts thus.}\
\par}
\end{samepage}

{\PaliColumn{796.}}\\

\begin{samepage}
{\PaliColumn{na brāhmaṇo aññato suddhimāha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{A brahmin does not speak of purity by another,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhe sute sīlavate mute vā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{by the seen and the heard, by good behavior and observances, by the sensed.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{puññe ca pāpe ca anūpalitto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Untainted by merit and by evil, he has discarded}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{attañjaho nayidha pakubbamāno.}}\\
{\setstretch{.7}
\quad\EnglishColumn{what was taken up without creating anything here.}\
\par}
\end{samepage}

{\PaliColumn{797.}}\\

\begin{samepage}
{\PaliColumn{purimaṃ pahāya aparaṃ sitāse,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having abandoned the former, attached to another,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ejānugā te na taranti saṅgaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{carried along by impulse, they do not cross the tie.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{te uggahāyanti nirassajanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{They grab hold and let go, like a monkey}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kapīva sākhaṃ pamuñcaṃ gahāyaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{grasping and letting go of a branch.}\
\par}
\end{samepage}

{\PaliColumn{798.}}\\

\begin{samepage}
{\PaliColumn{sayaṃ samādāya vatāni jantu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having undertaken observances by himself,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{uccāvacaṃ gacchati saññasatto;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a person goes up and down, attached to perception.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vidvā ca vedehi samecca dhammaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But having realized the Dhamma with knowledge,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na uccāvacaṃ gacchati bhūripañño.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the wise one, broad in wisdom, does not go up and down.}\
\par}
\end{samepage}

{\PaliColumn{799.}}\\

\begin{samepage}
{\PaliColumn{sa sabbadhammesu visenibhūto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One who is remote from all phenomena,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yaṃ kiñci diṭṭhaṃ va sutaṃ mutaṃ vā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{from whatever is seen, heard, or sensed —}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tameva dassiṃ vivaṭaṃ carantaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{how could anyone here in the world categorize him,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kenīdha lokasmi vikappayeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{that seer behaving openly?}\
\par}
\end{samepage}

{\PaliColumn{800.}}\\

\begin{samepage}
{\PaliColumn{na kappayanti na purekkharonti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{They do not construct, they have no preferences,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{accantasuddhīti na te vadanti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{they do not assert: “[This is] ultimate purity.”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ādānaganthaṃ gathitaṃ visajja,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having loosened the knot of grasping that had been tied,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{āsaṃ na kubbanti kuhiñci loke.}}\\
{\setstretch{.7}
\quad\EnglishColumn{they do not form desires for anything in the world.}\
\par}
\end{samepage}

{\PaliColumn{801.}}\\

\begin{samepage}
{\PaliColumn{sīmātigo brāhmaṇo tassa natthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{For a brahmin who has transcended the boundary,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñatvā va disvā va samuggahītaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{who has known and seen, nothing is tightly grasped.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na rāgarāgī na virāgaratto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not excited by lust or attached to dispassion,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tassīdha natthī paramuggahītanti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he does not grasp anything here as supreme.}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{suddhaṭṭhakasuttaṃ catutthaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{5. paramaṭṭhakasuttaṃ (KN 5.43)}}\\
{\setstretch{.7}
\quad\EnglishColumn{The Octad on the Supreme}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{802.}}\\

\begin{samepage}
{\PaliColumn{paramanti diṭṭhīsu paribbasāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Settling [on his own] as supreme among views,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yaduttari kurute jantu loke;}}\\
{\setstretch{.7}
\quad\EnglishColumn{whatever a person esteems as best in the world,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hīnāti aññe tato sabbamāha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{[in comparison] he says all others are “inferior”:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā vivādāni avītivatto.}}\\
{\setstretch{.7}
\quad\EnglishColumn{therefore he has not transcended disputes.}\
\par}
\end{samepage}

{\PaliColumn{803.}}\\

\begin{samepage}
{\PaliColumn{yadattanī passati ānisaṃsaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Whatever benefit one sees for oneself}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhe sute sīlavate mute vā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{in the seen, the heard, the sensed, or in good behavior and observances,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tadeva so tattha samuggahāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having grasped hold of that alone,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nihīnato passati sabbamaññaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one regards all else as inferior.}\
\par}
\end{samepage}

{\PaliColumn{804.}}\\

\begin{samepage}
{\PaliColumn{taṃ vāpi ganthaṃ kusalā vadanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{The skilled speak of that as a knot}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yaṃ nissito passati hīnamaññaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{when one is attached and regards others as inferior.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā hi diṭṭhaṃ va sutaṃ mutaṃ vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Therefore a bhikkhu should not be attached to the seen,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sīlabbataṃ bhikkhu na nissayeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{to the heard or sensed, or to good behavior and observances.}\
\par}
\end{samepage}

{\PaliColumn{805.}}\\

\begin{samepage}
{\PaliColumn{diṭṭhimpi lokasmiṃ na kappayeyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Nor should one construct any view in the world}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñāṇena vā sīlavatena vāpi;}}\\
{\setstretch{.7}
\quad\EnglishColumn{by means of knowledge or good behavior and observances.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{samoti attānamanūpaneyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should not take oneself as “equal”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hīno na maññetha visesi vāpi.}}\\
{\setstretch{.7}
\quad\EnglishColumn{or think of oneself as “inferior” or “superior.”}\
\par}
\end{samepage}

{\PaliColumn{806.}}\\

\begin{samepage}
{\PaliColumn{attaṃ pahāya anupādiyāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having abandoned what is taken up, not clinging,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñāṇepi so nissayaṃ no karoti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{one does not create a dependency even on knowledge.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa ve viyattesu na vaggasārī,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not taking sides among those who are divided,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhimpi so na pacceti kiñci.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one does not fall back on any view at all.}\
\par}
\end{samepage}

{\PaliColumn{807.}}\\

\begin{samepage}
{\PaliColumn{yassūbhayante paṇidhīdha natthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{For one who has no wish here for either end,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhavābhavāya idha vā huraṃ vā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{for various states of existence here or beyond,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nivesanā tassa na santi keci,}}\\
{\setstretch{.7}
\quad\EnglishColumn{there are no places of residence at all}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dhammesu niccheyya samuggahītaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{grasped after deciding among teachings.}\
\par}
\end{samepage}

{\PaliColumn{808.}}\\

\begin{samepage}
{\PaliColumn{tassīdha diṭṭhe va sute mute vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not even a subtle notion is formulated by him}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pakappitā natthi aṇūpi saññā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{about what is seen, heard, or sensed here.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{taṃ brāhmaṇaṃ diṭṭhimanādiyānaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{How could anyone here in the world categorize him,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kenīdha lokasmiṃ vikappayeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{that brahmin who does not cling to any view?}\
\par}
\end{samepage}

{\PaliColumn{809.}}\\

\begin{samepage}
{\PaliColumn{na kappayanti na purekkharonti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{They do not construct, they have no preferences;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dhammāpi tesaṃ na paṭicchitāse;}}\\
{\setstretch{.7}
\quad\EnglishColumn{even the teachings are not embraced by them.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na brāhmaṇo sīlavatena neyyo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{A brahmin cannot be led by good behavior and observances;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pāraṅgato na pacceti tādīti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the impartial one, gone beyond, does not fall back.}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{paramaṭṭhakasuttaṃ pañcamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{6. jarāsuttaṃ (KN 5.44)}}\\
{\setstretch{.7}
\quad\EnglishColumn{Old Age}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{810.}}\\

\begin{samepage}
{\PaliColumn{appaṃ vata jīvitaṃ idaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Short, alas, is this life;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{oraṃ vassasatāpi miyyati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{one even dies before a hundred years.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yo cepi aticca jīvati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Even if one lives longer,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atha kho so jarasāpi miyyati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one still dies because of old age.}\
\par}
\end{samepage}

{\PaliColumn{811.}}\\

\begin{samepage}
{\PaliColumn{socanti janā mamāyite,}}\\
{\setstretch{.7}
\quad\EnglishColumn{People sorrow over things taken as “mine,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na hi santi niccā pariggahā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{for there are no permanent possessions.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vinābhāvasantamevidaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having seen that there is separation,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{iti disvā nāgāramāvase.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not live the home life.}\
\par}
\end{samepage}

{\PaliColumn{812.}}\\

\begin{samepage}
{\PaliColumn{maraṇenapi taṃ pahīyati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Whatever a person conceives thus, “This is mine,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yaṃ puriso mamidanti maññati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{that too is abandoned at death.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etampi viditvā paṇḍito,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having understood this, a wise one, my follower,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na mamattāya nametha māmako.}}\\
{\setstretch{.7}
\quad\EnglishColumn{should not incline to take things as “mine.”}\
\par}
\end{samepage}

{\PaliColumn{813.}}\\

\begin{samepage}
{\PaliColumn{supinena yathāpi saṅgataṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Just as, on awakening, a person does not see}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paṭibuddho puriso na passati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{whatever was encountered in a dream,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evampi piyāyitaṃ janaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{so too one does not see a beloved person}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{petaṃ kālakataṃ na passati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{who has died and passed away.}\
\par}
\end{samepage}

{\PaliColumn{814.}}\\

\begin{samepage}
{\PaliColumn{diṭṭhāpi sutāpi te janā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Although those people were seen and heard,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yesaṃ nāmamidaṃ pavuccati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and were known by such and such a name,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāmaṃyevāvasissati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{when a person has passed away,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{akkheyyaṃ petassa jantuno.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the name alone remains to be uttered.}\
\par}
\end{samepage}

{\PaliColumn{815.}}\\

\begin{samepage}
{\PaliColumn{sokapparidevamaccharaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Those who are greedy for personal assets}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na jahanti giddhā mamāyite;}}\\
{\setstretch{.7}
\quad\EnglishColumn{do not escape sorrow, lamentation, and miserliness.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā munayo pariggahaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Therefore the munis, seers of security,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hitvā acariṃsu khemadassino.}}\\
{\setstretch{.7}
\quad\EnglishColumn{wandered having abandoned possessions.}\
\par}
\end{samepage}

{\PaliColumn{816.}}\\

\begin{samepage}
{\PaliColumn{patilīnacarassa bhikkhuno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{When a bhikkhu lives withdrawn,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhajamānassa vivittamāsanaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{resorting to a secluded seat,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sāmaggiyamāhu tassa taṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{they say it is appropriate for him}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yo attānaṃ bhavane na dassaye.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not to show himself in a fixed dwelling.}\
\par}
\end{samepage}

{\PaliColumn{817.}}\\

\begin{samepage}
{\PaliColumn{sabbattha munī anissito,}}\\
{\setstretch{.7}
\quad\EnglishColumn{The muni is not dependent on anything;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na piyaṃ kubbati nopi appiyaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he takes nothing as pleasing, nothing as displeasing.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmiṃ paridevamaccharaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Lamentation and miserliness do not stick to him}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paṇṇe vāri yathā na limpati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{as water does not stick to a leaf.}\
\par}
\end{samepage}

{\PaliColumn{818.}}\\

\begin{samepage}
{\PaliColumn{udabindu yathāpi pokkhare,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Just as a water drop does not stick to a lotus leaf,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{padume vāri yathā na limpati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{or water to the lotus flower,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evaṃ muni nopalimpati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{so the muni does not cling to anything}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yadidaṃ diṭṭhasutaṃ mutesu vā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{among the seen, heard, or sensed.}\
\par}
\end{samepage}

{\PaliColumn{819.}}\\

\begin{samepage}
{\PaliColumn{dhono na hi tena maññati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One cleansed does not thereby conceive}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yadidaṃ diṭṭhasutaṃ mutesu vā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{things seen, heard, or sensed.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāññena visuddhimicchati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He does not wish for purification by another,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na hi so rajjati no virajjatīti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for he becomes neither passionate nor dispassionate.}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{jarāsuttaṃ chaṭṭhaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{7. tissametteyyasuttaṃ (KN 5.45)}}\\
{\setstretch{.7}
\quad\EnglishColumn{Tissa Metteyya}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{820.}}\\

\begin{samepage}
{\PaliColumn{“methunamanuyuttassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“For one who indulges in sexual intercourse,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(iccāyasmā tisso metteyyo)}}\\
{\setstretch{.7}
\quad\EnglishColumn{(said the Venerable Tissa Metteyya),}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vighātaṃ brūhi mārisa;}}\\
{\setstretch{.7}
\quad\EnglishColumn{“tell me, dear sir, of the distress.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sutvāna tava sāsanaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having heard your teaching,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{viveke sikkhissāmase.}}\\
{\setstretch{.7}
\quad\EnglishColumn{we will train in seclusion.”}\
\par}
\end{samepage}

{\PaliColumn{821.}}\\

\begin{samepage}
{\PaliColumn{“methunamanuyuttassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“When one indulges in sexual intercourse,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(metteyyāti bhagavā)}}\\
{\setstretch{.7}
\quad\EnglishColumn{(Metteyya,” the Blessed One said),}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mussate vāpi sāsanaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{“even the teaching itself is forgotten,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{micchā ca paṭipajjati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{and he practices wrongly:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etaṃ tasmiṃ anāriyaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{this is ignoble in him.}\
\par}
\end{samepage}

{\PaliColumn{822.}}\\

\begin{samepage}
{\PaliColumn{“eko pubbe caritvāna,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One who previously lived alone}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{methunaṃ yo nisevati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{but then resorts to sexual intercourse}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yānaṃ bhantaṃ va taṃ loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{is like a vehicle that has gone astray:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hīnamāhu puthujjanaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{in the world they call him a low worldling.}\
\par}
\end{samepage}

{\PaliColumn{823.}}\\

\begin{samepage}
{\PaliColumn{“yaso kitti ca yā pubbe,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Whatever fame and acclaim he previously enjoyed,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hāyate vāpi tassa sā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{these fall away from him.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etampi disvā sikkhetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having seen this, one should train}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{methunaṃ vippahātave.}}\\
{\setstretch{.7}
\quad\EnglishColumn{to abandon sexual intercourse.}\
\par}
\end{samepage}

{\PaliColumn{824.}}\\

\begin{samepage}
{\PaliColumn{“saṅkappehi pareto so,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Oppressed by his thoughts,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kapaṇo viya jhāyati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he broods like a poor wretch.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sutvā paresaṃ nigghosaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having heard the reprimand of others,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{maṅku hoti tathāvidho.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one such as this is humiliated.}\
\par}
\end{samepage}

{\PaliColumn{825.}}\\

\begin{samepage}
{\PaliColumn{“atha satthāni kurute,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Then he prepares weapons [of defense]}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paravādehi codito;}}\\
{\setstretch{.7}
\quad\EnglishColumn{when reproached by the words of others.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{esa khvassa mahāgedho,}}\\
{\setstretch{.7}
\quad\EnglishColumn{This is a great thicket for him,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mosavajjaṃ pagāhati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{that he sinks into false speech.}\
\par}
\end{samepage}

{\PaliColumn{826.}}\\

\begin{samepage}
{\PaliColumn{“paṇḍitoti samaññāto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He was known as a wise man}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ekacariyaṃ adhiṭṭhito;}}\\
{\setstretch{.7}
\quad\EnglishColumn{when resolved on living alone,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{athāpi methune yutto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{but when he engaged in sexual intercourse,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mandova parikissati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{like a dullard he is afflicted.}\
\par}
\end{samepage}

{\PaliColumn{827.}}\\

\begin{samepage}
{\PaliColumn{“etamādīnavaṃ ñatvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having known this danger}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{muni pubbāpare idha;}}\\
{\setstretch{.7}
\quad\EnglishColumn{here in the before and after,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ekacariyaṃ daḷhaṃ kayirā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{a muni should resolutely live alone;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na nisevetha methunaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should not resort to sexual intercourse.}\
\par}
\end{samepage}

{\PaliColumn{828.}}\\

\begin{samepage}
{\PaliColumn{“vivekaññeva sikkhetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One should train just in seclusion;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etadariyānamuttamaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{this is supreme among the noble ones.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na tena seṭṭho maññetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should not conceive oneself best because of this:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa ve nibbānasantike.}}\\
{\setstretch{.7}
\quad\EnglishColumn{then one is indeed close to nibbāna.}\
\par}
\end{samepage}

{\PaliColumn{829.}}\\

\begin{samepage}
{\PaliColumn{“rittassa munino carato,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“When the muni lives void,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāmesu anapekkhino;}}\\
{\setstretch{.7}
\quad\EnglishColumn{indifferent to sensual pleasures,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{oghatiṇṇassa pihayanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{people tied to sensual pleasures envy him}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāmesu gadhitā pajā”ti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{as one who has crossed the flood.”}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{tissametteyyasuttaṃ sattamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{8. pasūrasuttaṃ (KN 5.46)}}\\
{\setstretch{.7}
\quad\EnglishColumn{Pasūra}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{830.}}\\

\begin{samepage}
{\PaliColumn{idheva suddhi iti vādayanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“‘Here alone is purity,’ they declare;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāññesu dhammesu visuddhimāhu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{they say there is no purification in other teachings.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yaṃ nissitā tattha subhaṃ vadānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Declaring whatever they depend on to be excellent,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paccekasaccesu puthū niviṭṭhā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{many are entrenched in separate truths.}\
\par}
\end{samepage}

{\PaliColumn{831.}}\\

\begin{samepage}
{\PaliColumn{te vādakāmā parisaṃ vigayha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Desiring debate, having plunged into the assembly,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bālaṃ dahantī mithu aññamaññaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{in pairs they accuse one another of being a fool.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vadanti te aññasitā kathojjaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Those of different convictions assert their argument,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pasaṃsakāmā kusalā vadānā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{desiring praise, claiming to be skilled.}\
\par}
\end{samepage}

{\PaliColumn{832.}}\\

\begin{samepage}
{\PaliColumn{yutto kathāyaṃ parisāya majjhe,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Keen on speaking in the midst of the assembly,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pasaṃsamicchaṃ vinighāti hoti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{wishing for praise, he becomes anxious.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{apāhatasmiṃ pana maṅku hoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{If his assertion is rejected, he feels humiliated.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nindāya so kuppati randhamesī.}}\\
{\setstretch{.7}
\quad\EnglishColumn{Upset by blame, he seeks a flaw [in his opponent].}\
\par}
\end{samepage}

{\PaliColumn{833.}}\\

\begin{samepage}
{\PaliColumn{yamassa vādaṃ parihīnamāhu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“If they say his assertion is defective,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{apāhataṃ pañhavimaṃsakāse;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and the judges consider it rejected,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paridevati socati hīnavādo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{the loser laments and sorrows,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{upaccagā manti anutthunāti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he moans: ‘He surpassed me.’}\
\par}
\end{samepage}

{\PaliColumn{834.}}\\

\begin{samepage}
{\PaliColumn{ete vivādā samaṇesu jātā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“These disputes have arisen among ascetics;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etesu ugghāti nighāti hoti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{in relation to them, one becomes elated and dejected.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etampi disvā virame kathojjaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having seen this, too, one should desist from arguments,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na haññadatthatthipasaṃsalābhā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for there is no other benefit than praise and gain.}\
\par}
\end{samepage}

{\PaliColumn{835.}}\\

\begin{samepage}
{\PaliColumn{pasaṃsito vā pana tattha hoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“If, however, he is praised there,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{akkhāya vādaṃ parisāya majjhe;}}\\
{\setstretch{.7}
\quad\EnglishColumn{having declared his doctrine in the midst of the assembly,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{so hassatī uṇṇamatī ca tena,}}\\
{\setstretch{.7}
\quad\EnglishColumn{he is thrilled by this and swells with pride,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pappuyya tamatthaṃ yathā mano ahu.}}\\
{\setstretch{.7}
\quad\EnglishColumn{having obtained the benefit that accords with his wish.}\
\par}
\end{samepage}

{\PaliColumn{836.}}\\

\begin{samepage}
{\PaliColumn{yā uṇṇatī sāssa vighātabhūmi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Though his pride is the ground of distress,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mānātimānaṃ vadate paneso;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he yet speaks from conceit and arrogance.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etampi disvā na vivādayetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having seen this, too, one should not dispute;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na hi tena suddhiṃ kusalā vadanti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for the skilled say this does not bring purity.}\
\par}
\end{samepage}

{\PaliColumn{837.}}\\

\begin{samepage}
{\PaliColumn{sūro yathā rājakhādāya puṭṭho,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Just as a hero, nourished on the king’s food,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{abhigajjameti paṭisūramicchaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{comes thundering, desiring an opponent,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yeneva so tena palehi sūra,}}\\
{\setstretch{.7}
\quad\EnglishColumn{rush off in his direction, Sūra}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pubbeva natthi yadidaṃ yudhāya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{already in the past, there was no scope for a fight.}\
\par}
\end{samepage}

{\PaliColumn{838.}}\\

\begin{samepage}
{\PaliColumn{ye diṭṭhimuggayha vivādayanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“When they grasp a view and argue,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{idameva saccanti ca vādayanti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and assert, ‘This alone is true,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{te tvaṃ vadassū na hi tedha atthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{tell them: ‘When a debate has arisen}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vādamhi jāte paṭisenikattā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{there is no opponent for you here.’}\
\par}
\end{samepage}

{\PaliColumn{839.}}\\

\begin{samepage}
{\PaliColumn{visenikatvā pana ye caranti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“But among those who live remote,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhīhi diṭṭhiṃ avirujjhamānā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{who do not oppose a view with other views,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tesu tvaṃ kiṃ labhetho pasūra,}}\\
{\setstretch{.7}
\quad\EnglishColumn{what will you obtain, Pasūra, from those}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yesīdha natthī paramuggahītaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{who grasp nothing here as supreme?}\
\par}
\end{samepage}

{\PaliColumn{840.}}\\

\begin{samepage}
{\PaliColumn{atha tvaṃ pavitakkamāgamā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“But now you have come brooding,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{manasā diṭṭhigatāni cintayanto;}}\\
{\setstretch{.7}
\quad\EnglishColumn{devising speculative views in your mind.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dhonena yugaṃ samāgamā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{You have confronted one who is cleansed:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na hi tvaṃ sakkhasi sampayātaveti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{indeed, you won’t be able to succeed.”}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{pasūrasuttaṃ aṭṭhamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{9. māgaṇḍiyasuttaṃ (KN 5.47)}}\\
{\setstretch{.7}
\quad\EnglishColumn{Māgandiya}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{841.}}\\

\begin{samepage}
{\PaliColumn{“disvāna taṇhaṃ aratiṃ ragañca,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having seen Taṇhā, Aratī, and Ragā,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāhosi chando api methunasmiṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{I did not have any desire for sexual intercourse,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kimevidaṃ muttakarīsapuṇṇaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{so why [should I desire] this, full of urine and feces?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pādāpi naṃ samphusituṃ na icche”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{I would not wish to touch her even with my foot.”}\
\par}
\end{samepage}

{\PaliColumn{842.}}\\

\begin{samepage}
{\PaliColumn{“etādisaṃ ce ratanaṃ na icchasi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“If you do not wish a gem such as this,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāriṃ narindehi bahūhi patthitaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a woman desired by many rulers of men,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhigataṃ sīlavataṃ nu jīvitaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{what kind of view, behavior, observances, lifestyle,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhavūpapattiñca vadesi kīdisaṃ”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{existence, and rebirth do you assert?”}\
\par}
\end{samepage}

{\PaliColumn{843.}}\\

\begin{samepage}
{\PaliColumn{“idaṃ vadāmīti na tassa hoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having decided among teachings,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(māgaṇḍiyāti bhagavā)}}\\
{\setstretch{.7}
\quad\EnglishColumn{(Māgandiya,” said the Blessed One),}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dhammesu niccheyya samuggahītaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{“it does not occur to one, ‘I assert this,’ [about a view] tightly grasped.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{passañca diṭṭhīsu anuggahāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But seeing into views, not grasping any of them,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ajjhattasantiṃ pacinaṃ adassaṃ”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{investigating, I saw the peace within.”}\
\par}
\end{samepage}

{\PaliColumn{844.}}\\

\begin{samepage}
{\PaliColumn{“vinicchayā yāni pakappitāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Indeed, muni, you speak without grasping}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(iti māgaṇḍiyo )}}\\
{\setstretch{.7}
\quad\EnglishColumn{(said Māgandiya)}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{te ve munī brūsi anuggahāya;}}\\
{\setstretch{.7}
\quad\EnglishColumn{those judgments that have been formulated,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ajjhattasantīti yametamatthaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“As to that matter called ‘the peace within,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kathaṃ nu dhīrehi paveditaṃ taṃ”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{how is it proclaimed by the wise?”}\
\par}
\end{samepage}

{\PaliColumn{845.}}\\

\begin{samepage}
{\PaliColumn{“na diṭṭhiyā na sutiyā na ñāṇena,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Not by view, nor by learning, nor by knowledge,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(māgaṇḍiyāti bhagavā)}}\\
{\setstretch{.7}
\quad\EnglishColumn{(Māgandiya,” said the Blessed One),}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sīlabbatenāpi na suddhimāha;}}\\
{\setstretch{.7}
\quad\EnglishColumn{“nor do I speak of purity through good behavior and observances;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{adiṭṭhiyā assutiyā añāṇā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{but neither without view, without learning, without knowledge,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{asīlatā abbatā nopi tena.}}\\
{\setstretch{.7}
\quad\EnglishColumn{without good behavior, without observances — not in that way.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ete ca nissajja anuggahāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But having relinquished these, not grasping any of them,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{santo anissāya bhavaṃ na jappe”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{peaceful, not dependent, one should not hanker for existence.”}\
\par}
\end{samepage}

{\PaliColumn{846.}}\\

\begin{samepage}
{\PaliColumn{“no ce kira diṭṭhiyā na sutiyā na ñāṇena,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“If indeed it is not by view, by learning, nor by knowledge,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(iti māgaṇḍiyo)}}\\
{\setstretch{.7}
\quad\EnglishColumn{(said Māgandiya),}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sīlabbatenāpi na suddhimāha;}}\\
{\setstretch{.7}
\quad\EnglishColumn{“nor by good behavior and observances, that one speaks of purity;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{adiṭṭhiyā assutiyā añāṇā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{nor without view, without learning, without knowledge,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{asīlatā abbatā nopi tena.}}\\
{\setstretch{.7}
\quad\EnglishColumn{without good behavior and observances — not in that way,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{maññāmahaṃ momuhameva dhammaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{I think this is an utterly confused teaching;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhiyā eke paccenti suddhiṃ”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{some fall back on purity by means of view.”}\
\par}
\end{samepage}

{\PaliColumn{847.}}\\

\begin{samepage}
{\PaliColumn{“diṭṭhañca nissāya anupucchamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Asking repeatedly while dependent on a view,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(māgaṇḍiyāti bhagavā)}}\\
{\setstretch{.7}
\quad\EnglishColumn{(Māgandiya,” said the Blessed One),}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{samuggahītesu pamohamāgā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{“you have become baffled over things tightly grasped.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ito ca nāddakkhi aṇumpi saññaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But from this you have not gained even an inkling;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā tuvaṃ momuhato dahāsi.}}\\
{\setstretch{.7}
\quad\EnglishColumn{hence you consider it utterly confused.}\
\par}
\end{samepage}

{\PaliColumn{848.}}\\

\begin{samepage}
{\PaliColumn{“samo visesī uda vā nihīno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One who thinks himself equal, superior, or inferior}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yo maññatī so vivadetha tena;}}\\
{\setstretch{.7}
\quad\EnglishColumn{might engage in disputes because of this.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tīsu vidhāsu avikampamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not shaking among these three discriminations,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{samo visesīti na tassa hoti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he does not think ‘equal, superior.’}\
\par}
\end{samepage}

{\PaliColumn{849.}}\\

\begin{samepage}
{\PaliColumn{“saccanti so brāhmaṇo kiṃ vadeyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Why would that brahmin assert, ‘It’s true,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{musāti vā so vivadetha kena;}}\\
{\setstretch{.7}
\quad\EnglishColumn{or with whom would he dispute, ‘It’s false’?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yasmiṃ samaṃ visamaṃ vāpi natthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{When for him there is no ‘equal’ and ‘unequal,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa kena vādaṃ paṭisaṃyujeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{with whom would he engage in debate?}\
\par}
\end{samepage}

{\PaliColumn{850.}}\\

\begin{samepage}
{\PaliColumn{“okaṃ pahāya aniketasārī,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having left home to roam without abode,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{gāme akubbaṃ muni santhavāni;}}\\
{\setstretch{.7}
\quad\EnglishColumn{in the village the muni is intimate with none.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāmehi ritto apurekkharāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Void of sensual pleasures, without preferences,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kathaṃ na viggayha janena kayirā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he would not engage in contentious talk with people.}\
\par}
\end{samepage}

{\PaliColumn{851.}}\\

\begin{samepage}
{\PaliColumn{“yehi vivitto vicareyya loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“When he wanders detached from things in the world,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na tāni uggayha vadeyya nāgo;}}\\
{\setstretch{.7}
\quad\EnglishColumn{the nāga would not grasp and assert them.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{jalambujaṃ kaṇḍakaṃ vārijaṃ yathā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{As a thorny-stalked lotus, born in the water,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{jalena paṅkena canūpalittaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{is untainted by water and mud,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evaṃ munī santivādo agiddho,}}\\
{\setstretch{.7}
\quad\EnglishColumn{just so the muni, a proponent of peace, free of greed,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāme ca loke ca anūpalitto.}}\\
{\setstretch{.7}
\quad\EnglishColumn{is untainted by sensual pleasures and the world.}\
\par}
\end{samepage}

{\PaliColumn{852.}}\\

\begin{samepage}
{\PaliColumn{“na vedagū diṭṭhiyāyako na mutiyā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Because of a view or an opinion a Veda-master does not become conceited,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa mānameti na hi tammayo so;}}\\
{\setstretch{.7}
\quad\EnglishColumn{for he does not identify with them.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na kammunā nopi sutena neyyo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not led by kamma or by what is heard,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anūpanīto sa nivesanesu.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he is not drawn to any abodes.}\
\par}
\end{samepage}

{\PaliColumn{853.}}\\

\begin{samepage}
{\PaliColumn{“saññāvirattassa na santi ganthā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“For one detached from perception there are no knots;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paññāvimuttassa na santi mohā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{for one liberated by wisdom there are no delusions.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saññañca diṭṭhiñca ye aggahesuṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But those who have grasped perceptions and views}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{te ghaṭṭayantā vicaranti loke”ti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{wander in the world creating friction.”}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{māgaṇḍiyasuttaṃ navamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{10. purābhedasuttaṃ (KN 5.48)}}\\
{\setstretch{.7}
\quad\EnglishColumn{Before the Breakup}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{854.}}\\

\begin{samepage}
{\PaliColumn{“kathaṃdassī kathaṃsīlo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“How does he see, how does he behave,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{upasantoti vuccati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{the one who is said to be ‘peaceful’?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{taṃ me gotama pabrūhi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{When asked by me, O Gotama,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pucchito uttamaṃ naraṃ”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{describe the supreme person.”}\
\par}
\end{samepage}

{\PaliColumn{855.}}\\

\begin{samepage}
{\PaliColumn{“vītataṇho purā bhedā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Devoid of craving before the breakup,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(iti bhagavā) pubbamantamanissito;}}\\
{\setstretch{.7}
\quad\EnglishColumn{(said the Blessed One), “not dependent on the past,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vemajjhe nupasaṅkheyyo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{not to be reckoned in the middle,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tassa natthi purakkhataṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for him there are no preferences.}\
\par}
\end{samepage}

{\PaliColumn{856.}}\\

\begin{samepage}
{\PaliColumn{“akkodhano asantāsī,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He is without anger, unafraid,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{avikatthī akukkuco;}}\\
{\setstretch{.7}
\quad\EnglishColumn{not boastful, not regretful,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mantabhāṇī anuddhato,}}\\
{\setstretch{.7}
\quad\EnglishColumn{of speaking with reflection, not restless:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa ve vācāyato muni.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he is truly a muni controlled in speech.}\
\par}
\end{samepage}

{\PaliColumn{857.}}\\

\begin{samepage}
{\PaliColumn{“nirāsatti anāgate,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He is without attachment to the future;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atītaṃ nānusocati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he does not sorrow over the past.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vivekadassī phassesu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{As a seer of seclusion in the midst of contacts,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhīsu ca na nīyati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he is not led astray among views.}\
\par}
\end{samepage}

{\PaliColumn{858.}}\\

\begin{samepage}
{\PaliColumn{“patilīno akuhako,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Withdrawn, not a schemer,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{apihālu amaccharī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{without longing, not miserly,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{appagabbho ajeguccho,}}\\
{\setstretch{.7}
\quad\EnglishColumn{courteous, not [morally] repulsive,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pesuṇeyye ca no yuto.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not intent on slander.}\
\par}
\end{samepage}

{\PaliColumn{859.}}\\

\begin{samepage}
{\PaliColumn{“sātiyesu anassāvī,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Not swept up by enjoyments,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atimāne ca no yuto;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and not swollen with arrogance;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saṇho ca paṭibhānavā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{gentle, gifted with ingenuity,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na saddho na virajjati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not credulous, not growing dispassionate.}\
\par}
\end{samepage}

{\PaliColumn{860.}}\\

\begin{samepage}
{\PaliColumn{“lābhakamyā na sikkhati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He does not train from a desire for gain,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{alābhe ca na kuppati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{nor is he irritated over lack of gain.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aviruddho ca taṇhāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not hostile, because of craving}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rasesu nānugijjhati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he does not hanker after tastes.}\
\par}
\end{samepage}

{\PaliColumn{861.}}\\

\begin{samepage}
{\PaliColumn{“upekkhako sadā sato,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Equanimous, ever mindful,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na loke maññate samaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{in the world he does not conceive himself}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na visesī na nīceyyo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{to be equal, or superior, or inferior:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tassa no santi ussadā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for him there are no swellings.}\
\par}
\end{samepage}

{\PaliColumn{862.}}\\

\begin{samepage}
{\PaliColumn{“yassa nissayanā natthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He has no dependencies —}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñatvā dhammaṃ anissito;}}\\
{\setstretch{.7}
\quad\EnglishColumn{having known the Dhamma, he is independent.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhavāya vibhavāya vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{No craving is found in him}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{taṇhā yassa na vijjati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for existence or nonexistence.}\
\par}
\end{samepage}

{\PaliColumn{863.}}\\

\begin{samepage}
{\PaliColumn{“taṃ brūmi upasantoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“I call him ‘peaceful’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāmesu anapekkhinaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{who is indifferent to sensual pleasures.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ganthā tassa na vijjanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{In him no knots are found;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atarī so visattikaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he has crossed over attachment.}\
\par}
\end{samepage}

{\PaliColumn{864.}}\\

\begin{samepage}
{\PaliColumn{“na tassa puttā pasavo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He has no sons or cattle,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{khettaṃ vatthuñca vijjati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{nor does he possess fields or land.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{attā vāpi nirattā vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{In him there is nothing to be found}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na tasmiṃ upalabbhati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{as either taken up or rejected.}\
\par}
\end{samepage}

{\PaliColumn{865.}}\\

\begin{samepage}
{\PaliColumn{“yena naṃ vajjuṃ puthujjanā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“That by which they might speak of him —}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atho samaṇabrāhmaṇā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{worldlings as well as ascetics and brahmins —}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{taṃ tassa apurakkhataṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{is not esteemed by him;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā vādesu nejati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{therefore he is not stirred up by words.}\
\par}
\end{samepage}

{\PaliColumn{866.}}\\

\begin{samepage}
{\PaliColumn{“vītagedho amaccharī,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Devoid of greed, without miserliness,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na ussesu vadate muni;}}\\
{\setstretch{.7}
\quad\EnglishColumn{the muni does not speak [of himself]}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na samesu na omesu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{as among superiors, or equals, or inferiors.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kappaṃ neti akappiyo.}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not given to mental construction, he does not enter upon mental constructs.}\
\par}
\end{samepage}

{\PaliColumn{867.}}\\

\begin{samepage}
{\PaliColumn{“yassa loke sakaṃ natthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One who takes nothing in the world as his own,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{asatā ca na socati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and who does not sorrow over what is absent,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dhammesu ca na gacchati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{who does not enter upon things:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa ve santoti vuccatī”ti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he is truly said to be ‘peaceful.’ ”}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{purābhedasuttaṃ dasamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{11. kalahavivādasuttaṃ (KN 5.49)}}\\
{\setstretch{.7}
\quad\EnglishColumn{Quarrels and Disputes}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{868.}}\\

\begin{samepage}
{\PaliColumn{“kutopahūtā kalahā vivādā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“From where do quarrels and disputes arise,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paridevasokā sahamaccharā ca;}}\\
{\setstretch{.7}
\quad\EnglishColumn{lamentation, sorrow, and miserliness?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mānātimānā sahapesuṇā ca,}}\\
{\setstretch{.7}
\quad\EnglishColumn{From where do conceit and arrogance arise}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kutopahūtā te tadiṅgha brūhi”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{along with slander? Please tell me this.”}\
\par}
\end{samepage}

{\PaliColumn{869.}}\\

\begin{samepage}
{\PaliColumn{“piyappahūtā kalahā vivādā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Quarrels and disputes arise from what is pleasing,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paridevasokā sahamaccharā ca;}}\\
{\setstretch{.7}
\quad\EnglishColumn{as do lamentation, sorrow, and miserliness,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mānātimānā sahapesuṇā ca,}}\\
{\setstretch{.7}
\quad\EnglishColumn{conceit and arrogance along with slander.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{maccherayuttā kalahā vivādā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{Quarrels and disputes are connected with miserliness,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vivādajātesu ca pesuṇāni”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{and slanders occur when disputes arise.”200}\
\par}
\end{samepage}

{\PaliColumn{870.}}\\

\begin{samepage}
{\PaliColumn{“piyā su lokasmiṃ kutonidānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“From what do pleasing things in the world originate,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ye cāpi lobhā vicaranti loke;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and those states of greed that spread through the world?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{āsā ca niṭṭhā ca kutonidānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{From what do longing and fulfillment originate,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ye samparāyāya narassa honti”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{which a person has about the future?”}\
\par}
\end{samepage}

{\PaliColumn{871.}}\\

\begin{samepage}
{\PaliColumn{“chandānidānāni piyāni loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Pleasing things in the world originate from desire,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ye cāpi lobhā vicaranti loke;}}\\
{\setstretch{.7}
\quad\EnglishColumn{as do those states of greed that spread through the world.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{āsā ca niṭṭhā ca itonidānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{From this originate the longing and fulfillment}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ye samparāyāya narassa honti”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{that a person has about the future.”}\
\par}
\end{samepage}

{\PaliColumn{872.}}\\

\begin{samepage}
{\PaliColumn{“chando nu lokasmiṃ kutonidāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“From what in the world does desire originate?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vinicchayā cāpi kutopahūtā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{And from what do judgments too arise,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kodho mosavajjañca kathaṃkathā ca,}}\\
{\setstretch{.7}
\quad\EnglishColumn{and anger, false speech, and perplexity,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ye vāpi dhammā samaṇena vuttā”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{and those [other] things the Ascetic has mentioned?”}\
\par}
\end{samepage}

{\PaliColumn{873.}}\\

\begin{samepage}
{\PaliColumn{“sātaṃ asātanti yamāhu loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Desire originates based on what they say}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tamūpanissāya pahoti chando;}}\\
{\setstretch{.7}
\quad\EnglishColumn{is ‘pleasant’ or ‘unpleasant’ in the world.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rūpesu disvā vibhavaṃ bhavañca,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having seen the vanishing and coming-to-be of forms,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vinicchayaṃ kubbati jantu loke.}}\\
{\setstretch{.7}
\quad\EnglishColumn{a person forms a judgment in the world.}\
\par}
\end{samepage}

{\PaliColumn{874.}}\\

\begin{samepage}
{\PaliColumn{“kodho mosavajjañca kathaṃkathā ca,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Anger, false speech, and perplexity:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etepi dhammā dvayameva sante;}}\\
{\setstretch{.7}
\quad\EnglishColumn{these things, too, arise when that dyad exists.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kathaṃkathī ñāṇapathāya sikkhe,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One perplexed should train on the path of knowledge;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñatvā pavuttā samaṇena dhammā”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{having known, the Ascetic stated these things.”}\
\par}
\end{samepage}

{\PaliColumn{875.}}\\

\begin{samepage}
{\PaliColumn{“sātaṃ asātañca kutonidānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“From what do the pleasant and unpleasant originate?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kismiṃ asante na bhavanti hete;}}\\
{\setstretch{.7}
\quad\EnglishColumn{When what does not exist do these not come to be?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vibhavaṃ bhavañcāpi yametamatthaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{As to this matter of vanishing and coming-to-be,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etaṃ me pabrūhi yatonidānaṃ”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{tell me from what it originates.”}\
\par}
\end{samepage}

{\PaliColumn{876.}}\\

\begin{samepage}
{\PaliColumn{“phassanidānaṃ sātaṃ asātaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“The pleasant and unpleasant originate from contact;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{phasse asante na bhavanti hete;}}\\
{\setstretch{.7}
\quad\EnglishColumn{when contact does not exist, these do not come to be.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vibhavaṃ bhavañcāpi yametamatthaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{As to this matter of vanishing and coming-to-be,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etaṃ te pabrūmi itonidānaṃ”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{I tell you that it originates from this.”}\
\par}
\end{samepage}

{\PaliColumn{877.}}\\

\begin{samepage}
{\PaliColumn{“phasso nu lokasmi kutonidāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“From what in the world does contact originate?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pariggahā cāpi kutopahūtā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{From what do possessions too arise?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kismiṃ asante na mamattamatthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{When what does not exist is there no taking as ‘mine’?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kismiṃ vibhūte na phusanti phassā”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{When what has vanished do contacts not touch one?”}\
\par}
\end{samepage}

{\PaliColumn{878.}}\\

\begin{samepage}
{\PaliColumn{“nāmañca rūpañca paṭicca phasso,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Contacts are dependent upon name and form;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{icchānidānāni pariggahāni;}}\\
{\setstretch{.7}
\quad\EnglishColumn{possessions are based on desire.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{icchāyasantyā na mamattamatthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{When desire does not exist, there is no taking as ‘mine.’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rūpe vibhūte na phusanti phassā”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{When form has vanished, contacts do not touch one.”}\
\par}
\end{samepage}

{\PaliColumn{879.}}\\

\begin{samepage}
{\PaliColumn{“kathaṃsametassa vibhoti rūpaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“How must one attain for form to vanish?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sukhaṃ dukhañcāpi kathaṃ vibhoti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{How do pleasure and pain also vanish?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etaṃ me pabrūhi yathā vibhoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Please tell me this, how they vanish.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{taṃ jāniyāmāti me mano ahu”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{We would like to know that — such is my thought.”}\
\par}
\end{samepage}

{\PaliColumn{880.}}\\

\begin{samepage}
{\PaliColumn{“na saññasaññī na visaññasaññī,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Not percipient through perception, not percipient through disturbed perception,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nopi asaññī na vibhūtasaññī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{not altogether without perception, not percipient of what has vanished:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evaṃsametassa vibhoti rūpaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{form vanishes for one who has so attained,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saññānidānā hi papañcasaṅkhā”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for concepts due to proliferation are based on perception.”}\
\par}
\end{samepage}

{\PaliColumn{881.}}\\

\begin{samepage}
{\PaliColumn{“yaṃ taṃ apucchimha akittayī no,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“You explained to us whatever we asked you.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aññaṃ taṃ pucchāma tadiṅgha brūhi;}}\\
{\setstretch{.7}
\quad\EnglishColumn{Let us now ask something else: please tell me this.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ettāvataggaṃ nu vadanti heke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Do some wise men here say that at this point}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yakkhassa suddhiṃ idha paṇḍitāse.}}\\
{\setstretch{.7}
\quad\EnglishColumn{this is the foremost purity of the spirit,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{udāhu aññampi vadanti etto.}}\\
{\setstretch{.7}
\quad\EnglishColumn{or do they speak of it as different from this?”203}\
\par}
\end{samepage}

{\PaliColumn{882.}}\\

\begin{samepage}
{\PaliColumn{“ettāvataggampi vadanti heke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Some wise men here say that at this point}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yakkhassa suddhiṃ idha paṇḍitāse;}}\\
{\setstretch{.7}
\quad\EnglishColumn{this is the foremost purity of the spirit.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tesaṃ paneke samayaṃ vadanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But204 some among them, claiming to be skilled,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anupādisese kusalā vadānā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{speak of an attainment without residue remaining.}\
\par}
\end{samepage}

{\PaliColumn{883.}}\\

\begin{samepage}
{\PaliColumn{“ete ca ñatvā upanissitāti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having known these to be ‘dependent,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñatvā munī nissaye so vimaṃsī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and having known the dependencies, the muni, the investigator,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñatvā vimutto na vivādameti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having known, liberated, does not enter disputes;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhavābhavāya na sameti dhīro”ti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the wise one does not come upon various states of existence.”}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{kalahavivādasuttaṃ ekādasamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{12. cūḷabyūhasuttaṃ VAR (KN 5.50)}}\\
{\setstretch{.7}
\quad\EnglishColumn{The Smaller Discourse on Deployment}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{884.}}\\

\begin{samepage}
{\PaliColumn{sakaṃsakaṃdiṭṭhiparibbasānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Settled each in his own view,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{viggayha nānā kusalā vadanti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{contending, the skilled make diverse assertions:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yo evaṃ jānāti sa vedi dhammaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{‘One who knows this has understood the Dhamma;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{idaṃ paṭikkosamakevalī so.}}\\
{\setstretch{.7}
\quad\EnglishColumn{rejecting this, one is not consummate.’}\
\par}
\end{samepage}

{\PaliColumn{885.}}\\

\begin{samepage}
{\PaliColumn{evampi viggayha vivādayanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having contended thus, they dispute}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bālo paro akkusaloti cāhu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and say: ‘The opponent is a fool, unskilled.’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sacco nu vādo katamo imesaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Which assertion among these is truthful,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbeva hīme kusalā vadānā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for they all claim to be skilled?”}\
\par}
\end{samepage}

{\PaliColumn{886.}}\\

\begin{samepage}
{\PaliColumn{parassa ce dhammamanānujānaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“If one who does not affirm the opponent’s doctrine}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bālomako hoti nihīnapañño;}}\\
{\setstretch{.7}
\quad\EnglishColumn{is thereby a fool, a brute, deficient in wisdom,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbeva bālā sunihīnapaññā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{all indeed are fools, utterly deficient in wisdom:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbevime diṭṭhiparibbasānā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for all these are settled in their views.}\
\par}
\end{samepage}

{\PaliColumn{887.}}\\

\begin{samepage}
{\PaliColumn{sandiṭṭhiyā ceva na vīvadātā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“But if by their own view they are cleansed,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saṃsuddhapaññā kusalā mutīmā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{pure in wisdom, skilled, intelligent,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na tesaṃ koci parihīnapañño,}}\\
{\setstretch{.7}
\quad\EnglishColumn{there is none among them defective in wisdom,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhī hi tesampi tathā samattā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for their views are similarly taken up.}\
\par}
\end{samepage}

{\PaliColumn{888.}}\\

\begin{samepage}
{\PaliColumn{na vāhametaṃ tathiyanti brūmi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“I do not say, ‘This is correct,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yamāhu bālā mithu aññamaññaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{as the fools in pairs say to one another.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sakaṃsakaṃdiṭṭhimakaṃsu saccaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{They all take their own view to be true;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā hi bāloti paraṃ dahanti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{therefore they consider the opponent a fool.”}\
\par}
\end{samepage}

{\PaliColumn{889.}}\\

\begin{samepage}
{\PaliColumn{yamāhu saccaṃ tathiyanti eke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“That which some say is ‘true, correct,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tamāhu aññe tucchaṃ musāti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{others say is ‘hollow, false.’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evampi vigayha vivādayanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having contended thus, they dispute.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kasmā na ekaṃ samaṇā vadanti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{Why don’t ascetics speak in unison?”}\
\par}
\end{samepage}

{\PaliColumn{890.}}\\

\begin{samepage}
{\PaliColumn{ekañhi saccaṃ na dutīyamatthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Truth, indeed, is one — there is no second;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yasmiṃ pajā no vivade pajānaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a person who understands this would not dispute.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nānā te saccāni sayaṃ thunanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{These proclaim their own diverse truths;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā na ekaṃ samaṇā vadanti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{therefore ascetics do not speak in unison.”}\
\par}
\end{samepage}

{\PaliColumn{891.}}\\

\begin{samepage}
{\PaliColumn{kasmā nu saccāni vadanti nānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“But why do they assert diverse truths,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pavādiyāse kusalā vadānā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{those proponents who claim to be skilled?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saccāni sutāni bahūni nānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Are those truths actually many and diverse,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{udāhu te takkamanussaranti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{or do they follow a line of reasoning?”}\
\par}
\end{samepage}

{\PaliColumn{892.}}\\

\begin{samepage}
{\PaliColumn{na heva saccāni bahūni nānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Truths surely are not many and diverse,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aññatra saññāya niccāni loke;}}\\
{\setstretch{.7}
\quad\EnglishColumn{except by [mistakenly] perceiving permanent things in the world.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{takkañca diṭṭhīsu pakappayitvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But having formulated reasoning about views,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saccaṃ musāti dvayadhammamāhu.}}\\
{\setstretch{.7}
\quad\EnglishColumn{they assert the dyad ‘true’ and ‘false.’}\
\par}
\end{samepage}

{\PaliColumn{893.}}\\

\begin{samepage}
{\PaliColumn{diṭṭhe sute sīlavate mute vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“The seen, heard, good behavior and observances,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ete ca nissāya vimānadassī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{the sensed — dependent on these, he shows disdain.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vinicchaye ṭhatvā pahassamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Based on a judgment, derisive,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bālo paro akkusaloti cāha.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he says: ‘The opponent is a fool, unskilled.’}\
\par}
\end{samepage}

{\PaliColumn{894.}}\\

\begin{samepage}
{\PaliColumn{yeneva bāloti paraṃ dahāti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“As he considers the opponent a fool,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tenātumānaṃ kusaloti cāha;}}\\
{\setstretch{.7}
\quad\EnglishColumn{on the same ground he describes himself as ‘skilled.’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sayamattanā so kusalo vadāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Of his own accord, claiming himself skilled,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aññaṃ vimāneti tadeva pāva.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he disdains the other, yet speaks in the same way.}\
\par}
\end{samepage}

{\PaliColumn{895.}}\\

\begin{samepage}
{\PaliColumn{atisāradiṭṭhiyāva so samatto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Inflated by that extremist view,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mānena matto paripuṇṇamānī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{intoxicated with conceit, thinking himself perfect,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sayameva sāmaṃ manasābhisitto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{on his own accord he has mentally anointed himself,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhī hi sā tassa tathā samattā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for that view of his is taken up in such a manner.}\
\par}
\end{samepage}

{\PaliColumn{896.}}\\

\begin{samepage}
{\PaliColumn{parassa ce hi vacasā nihīno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“If one is deficient because of the opponent’s word,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tumo sahā hoti nihīnapañño;}}\\
{\setstretch{.7}
\quad\EnglishColumn{the opponent himself is similarly deficient in wisdom.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atha ce sayaṃ vedagū hoti dhīro,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But if he is himself a master of knowledge, a wise man,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na koci bālo samaṇesu atthi.}}\\
{\setstretch{.7}
\quad\EnglishColumn{then there is no fool among ascetics.}\
\par}
\end{samepage}

{\PaliColumn{897.}}\\

\begin{samepage}
{\PaliColumn{aññaṃ ito yābhivadanti dhammaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Those who assert a teaching different from this}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aparaddhā suddhimakevalī te;}}\\
{\setstretch{.7}
\quad\EnglishColumn{have failed to reach purity and perfection:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evampi titthyā puthuso vadanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{thus the sectarians speak in separate ways,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sandiṭṭhirāgena hi tebhirattā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for they are attached to their own views.}\
\par}
\end{samepage}

{\PaliColumn{898.}}\\

\begin{samepage}
{\PaliColumn{idheva suddhi iti vādayanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“‘Here only is purity’ they assert;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāññesu dhammesu visuddhimāhu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{they say there is no purification in other teachings.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evampi titthyā puthuso niviṭṭhā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Thus, too, the sectarians, separately entrenched,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sakāyane tattha daḷhaṃ vadānā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{firmly assert their own way there.}\
\par}
\end{samepage}

{\PaliColumn{899.}}\\

\begin{samepage}
{\PaliColumn{sakāyane vāpi daḷhaṃ vadāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Asserting firmly his own way,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kamettha bāloti paraṃ daheyya;}}\\
{\setstretch{.7}
\quad\EnglishColumn{what opponent here could one consider a fool?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sayameva so medhagamāvaheyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He himself would just provoke strife}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paraṃ vadaṃ bālamasuddhidhammaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{by calling his opponent a fool of impure character.}\
\par}
\end{samepage}

{\PaliColumn{900.}}\\

\begin{samepage}
{\PaliColumn{vinicchaye ṭhatvā sayaṃ pamāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Based on a judgment, taking himself as the measure,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{uddhaṃ sa lokasmiṃ vivādameti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he enters upon further disputes in the world.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hitvāna sabbāni vinicchayāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having abandoned all judgments,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na medhagaṃ kubbati jantu loketi.}}\\
{\setstretch{.7}
\quad\EnglishColumn{a person does not create strife in the world.”}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{cūḷabyūhasuttaṃ dvādasamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{13. mahābyūhasuttaṃ (KN 5.51)}}\\
{\setstretch{.7}
\quad\EnglishColumn{The Greater Discourse on Deployment}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{901.}}\\

\begin{samepage}
{\PaliColumn{ye kecime diṭṭhiparibbasānā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Those who are settled in views,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{idameva saccanti vivādayanti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{who dispute, saying, ‘This alone is truth’:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbeva te nindamanvānayanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{do all of them receive only blame,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atho pasaṃsampi labhanti tattha.}}\\
{\setstretch{.7}
\quad\EnglishColumn{or do some there also win praise?”}\
\par}
\end{samepage}

{\PaliColumn{902.}}\\

\begin{samepage}
{\PaliColumn{appañhi etaṃ na alaṃ samāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“This [praise] is slight, not sufficient for peace;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{duve vivādassa phalāni brūmi;}}\\
{\setstretch{.7}
\quad\EnglishColumn{I say there are two fruits of disputes.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etampi disvā na vivādayetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having seen this too, one should not dispute,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{khemābhipassaṃ avivādabhūmiṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{seeing as security the stage of non-dispute.}\
\par}
\end{samepage}

{\PaliColumn{903.}}\\

\begin{samepage}
{\PaliColumn{yā kācimā sammutiyo puthujjā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Whatever commonplace opinions there are,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbāva etā na upeti vidvā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a wise person does not get involved in them.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anūpayo so upayaṃ kimeyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Why would one uninvolved become involved}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhe sute khantimakubbamāno.}}\\
{\setstretch{.7}
\quad\EnglishColumn{when he does not acquiesce in what is seen and heard?}\
\par}
\end{samepage}

{\PaliColumn{904.}}\\

\begin{samepage}
{\PaliColumn{sīluttamā saññamenāhu suddhiṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Those who take good behavior as supreme}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vataṃ samādāya upaṭṭhitāse;}}\\
{\setstretch{.7}
\quad\EnglishColumn{speak of purity through self-control.  Having taken up an observance, they settle on it,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{idheva sikkhema athassa suddhiṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{thinking: ‘Let’s train right here, then there would be purity.’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhavūpanītā kusalā vadānā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{Claiming to be skilled, they are led back into existence.}\
\par}
\end{samepage}

{\PaliColumn{905.}}\\

\begin{samepage}
{\PaliColumn{sace cuto sīlavatato hoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“If he has fallen away from good behavior and observances,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pavedhatī kamma virādhayitvā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he trembles because he has failed in his action.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pajappatī patthayatī ca suddhiṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He longs and yearns for purity here,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{satthāva hīno pavasaṃ gharamhā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{like one on a journey who has lost his caravan.}\
\par}
\end{samepage}

{\PaliColumn{906.}}\\

\begin{samepage}
{\PaliColumn{sīlabbataṃ vāpi pahāya sabbaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“But having abandoned all good behavior and observances,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kammañca sāvajjanavajjametaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and these deeds, blameworthy and blameless,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{suddhiṃ asuddhinti apatthayāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{not yearning for either purity or impurity,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{virato care santimanuggahāya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should live detached, not grasping peace.}\
\par}
\end{samepage}

{\PaliColumn{907.}}\\

\begin{samepage}
{\PaliColumn{tamūpanissāya jigucchitaṃ vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Dependent on austerity or scrupulousness,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{athavāpi diṭṭhaṃ va sutaṃ mutaṃ vā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{or on what is seen, heard, or sensed,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{uddhaṃsarā suddhimanutthunanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{running onward, they declare purity,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{avītataṇhāse bhavābhavesu.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not rid of craving for various states of existence.}\
\par}
\end{samepage}

{\PaliColumn{908.}}\\

\begin{samepage}
{\PaliColumn{patthayamānassa hi jappitāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“For one who yearns there are longings}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pavedhitaṃ vāpi pakappitesu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and trembling too over things mentally formulated;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{cutūpapāto idha yassa natthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{but for one here who has no passing away or rebirth,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa kena vedheyya kuhiṃva jappe.}}\\
{\setstretch{.7}
\quad\EnglishColumn{why would he tremble and for what would he long?”}\
\par}
\end{samepage}

{\PaliColumn{909.}}\\

\begin{samepage}
{\PaliColumn{yamāhu dhammaṃ paramanti eke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“The teaching that some here say is supreme,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tameva hīnanti panāhu aññe;}}\\
{\setstretch{.7}
\quad\EnglishColumn{others say that the same is inferior.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sacco nu vādo katamo imesaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Which assertion of theirs could be true}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbeva hīme kusalā vadānā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{when they all claim to be skilled?”}\
\par}
\end{samepage}

{\PaliColumn{910.}}\\

\begin{samepage}
{\PaliColumn{sakañhi dhammaṃ paripuṇṇamāhu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“They say their own teaching is complete,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aññassa dhammaṃ pana hīnamāhu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{but they say the other’s teaching is inferior.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evampi viggayha vivādayanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having quarreled thus, they go on disputing,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sakaṃ sakaṃ sammutimāhu saccaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{as they each say their own opinion is true.}\
\par}
\end{samepage}

{\PaliColumn{911.}}\\

\begin{samepage}
{\PaliColumn{parassa ce vambhayitena hīno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“If one is inferior when disparaged by an opponent,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na koci dhammesu visesi assa;}}\\
{\setstretch{.7}
\quad\EnglishColumn{no one would be distinguished among teachings.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{puthū hi aññassa vadanti dhammaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{For they each say the other’s teaching is inferior,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nihīnato samhi daḷhaṃ vadānā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{while firmly advocating their own.}\
\par}
\end{samepage}

{\PaliColumn{912.}}\\

\begin{samepage}
{\PaliColumn{saddhammapūjāpi nesaṃ tatheva,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“And their veneration of their own teaching,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yathā pasaṃsanti sakāyanāni;}}\\
{\setstretch{.7}
\quad\EnglishColumn{is done just as they praise their own ways.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbeva vādā tathiyā bhaveyyuṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{All their assertions would be truthful,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{suddhī hi nesaṃ paccattameva.}}\\
{\setstretch{.7}
\quad\EnglishColumn{since for them purity is exclusively their own.}\
\par}
\end{samepage}

{\PaliColumn{913.}}\\

\begin{samepage}
{\PaliColumn{na brāhmaṇassa paraneyyamatthi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“For a brahmin there is no being led by others,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dhammesu niccheyya samuggahītaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{no selecting among teachings and grasping tightly.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā vivādāni upātivatto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Therefore he has overcome disputes,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na hi seṭṭhato passati dhammamaññaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{since he does not regard another teaching as supreme.}\
\par}
\end{samepage}

{\PaliColumn{914.}}\\

\begin{samepage}
{\PaliColumn{jānāmi passāmi tatheva etaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Saying, ‘I know, I see, it is just like this,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{diṭṭhiyā eke paccenti suddhiṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{some here fall back on purity through a view.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{addakkhi ce kiñhi tumassa tena,}}\\
{\setstretch{.7}
\quad\EnglishColumn{If one has seen, what good is this to himself?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atisitvā aññena vadanti suddhiṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having gone too far, they assert purity by another.}\
\par}
\end{samepage}

{\PaliColumn{915.}}\\

\begin{samepage}
{\PaliColumn{passaṃ naro dakkhati nāmarūpaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Seeing, a person will see name-and-form;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{disvāna vā ñassati tānimeva;}}\\
{\setstretch{.7}
\quad\EnglishColumn{having seen, it is just these that he will know.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāmaṃ bahuṃ passatu appakaṃ vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Granted, let him see much or little,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na hi tena suddhiṃ kusalā vadanti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the skillful say purity is not won in that way.}\
\par}
\end{samepage}

{\PaliColumn{916.}}\\

\begin{samepage}
{\PaliColumn{nivissavādī na hi subbināyo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“A dogmatist is not easily disciplined,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pakappitaṃ diṭṭhi purekkharāno;}}\\
{\setstretch{.7}
\quad\EnglishColumn{one preferring a formulated view.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yaṃ nissito tattha subhaṃ vadāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Claiming that the good is found in what he depends on,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{suddhiṃvado tattha tathaddasā so.}}\\
{\setstretch{.7}
\quad\EnglishColumn{as a proponent of purity, he saw things that way there.}\
\par}
\end{samepage}

{\PaliColumn{917.}}\\

\begin{samepage}
{\PaliColumn{na brāhmaṇo kappamupeti saṅkhā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having comprehended, a brahmin does not take up mental constructs;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na diṭṭhisārī napi ñāṇabandhu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he is not a pursuer of views nor a kinsman of knowledge.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñatvā ca so sammutiyo puthujjā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having known the commonplace opinions,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{upekkhatī uggahaṇanti maññe.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he maintains equanimity while others grasp.}\
\par}
\end{samepage}

{\PaliColumn{918.}}\\

\begin{samepage}
{\PaliColumn{vissajja ganthāni munīdha loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having loosened the knots here in the world,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vivādajātesu na vaggasārī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{the muni does not take sides in arisen disputes.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{santo asantesu upekkhako so,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He is peaceful among the restless, equanimous,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anuggaho uggahaṇanti maññe.}}\\
{\setstretch{.7}
\quad\EnglishColumn{without grasping while others grasp.}\
\par}
\end{samepage}

{\PaliColumn{919.}}\\

\begin{samepage}
{\PaliColumn{pubbāsave hitvā nave akubbaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having abandoned past influxes, not creating new ones,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na chandagū nopi nivissavādī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he does not go along with desire, nor is he a dogmatist.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa vippamutto diṭṭhigatehi dhīro,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Released from speculative views, the wise person,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na limpati loke anattagarahī.}}\\
{\setstretch{.7}
\quad\EnglishColumn{free of self-reproach, is not tainted by the world.}\
\par}
\end{samepage}

{\PaliColumn{920.}}\\

\begin{samepage}
{\PaliColumn{sa sabbadhammesu visenibhūto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He is remote from all phenomena,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yaṃ kiñci diṭṭhaṃ va sutaṃ mutaṃ vā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{from whatever is seen, heard, or sensed.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa pannabhāro muni vippamutto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{With his burden dropped, released, the muni,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na kappiyo nūparato na patthiyoti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not given to mental construction, does not desist, does not yearn”}\
\par}
\end{samepage}

{\EnglishColumn{so said the Blessed One.}}\\

{\PaliColumn{mahābyūhasuttaṃ terasamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{14. tuvaṭakasuttaṃ (KN 5.52)}}\\
{\setstretch{.7}
\quad\EnglishColumn{Quickly}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{921.}}\\

\begin{samepage}
{\PaliColumn{“pucchāmi taṃ ādiccabandhu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“I ask you, Kinsman of the Sun, great rishi,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vivekaṃ santipadañca mahesi;}}\\
{\setstretch{.7}
\quad\EnglishColumn{about seclusion and the state of peace.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kathaṃ disvā nibbāti bhikkhu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{How having seen does a bhikkhu attain nibbāna,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anupādiyāno lokasmiṃ kiñci”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not clinging to anything in the world?”}\
\par}
\end{samepage}

{\PaliColumn{922.}}\\

\begin{samepage}
{\PaliColumn{“mūlaṃ papañcasaṅkhāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“By reflection, he should stop [the conceit] ‘I am,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(iti bhagavā)}}\\
{\setstretch{.7}
\quad\EnglishColumn{[the Blessed One said].}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mantā asmīti sabbamuparundhe;}}\\
{\setstretch{.7}
\quad\EnglishColumn{the entire root of concepts due to proliferation,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yā kāci taṇhā ajjhattaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Whatever cravings there may be internally,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tāsaṃ vinayā sadā sato sikkhe.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should always train mindfully for their removal.}\
\par}
\end{samepage}

{\PaliColumn{923.}}\\

\begin{samepage}
{\PaliColumn{“yaṃ kiñci dhammamabhijaññā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Whatever one might know,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ajjhattaṃ athavāpi bahiddhā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{whether internally or externally,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na tena thāmaṃ kubbetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not be obstinate on that account,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na hi sā nibbuti sataṃ vuttā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for that is not called quenching by the good.}\
\par}
\end{samepage}

{\PaliColumn{924.}}\\

\begin{samepage}
{\PaliColumn{“seyyo na tena maññeyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Because of this one should not think oneself better,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nīceyyo athavāpi sarikkho;}}\\
{\setstretch{.7}
\quad\EnglishColumn{nor should one consider oneself inferior or equal.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{phuṭṭho anekarūpehi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Being affected in various ways,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nātumānaṃ vikappayaṃ tiṭṭhe.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not persist in positioning oneself.}\
\par}
\end{samepage}

{\PaliColumn{925.}}\\

\begin{samepage}
{\PaliColumn{“ajjhattamevupasame,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“It is internally that he should achieve peace;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na aññato bhikkhu santimeseyya;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a bhikkhu should not seek peace through another.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ajjhattaṃ upasantassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{For one who is at peace within himself,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{natthi attā kuto nirattā vā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{there is nothing taken up, much less rejected.}\
\par}
\end{samepage}

{\PaliColumn{926.}}\\

\begin{samepage}
{\PaliColumn{“majjhe yathā samuddassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Just as in the middle of the ocean}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ūmi no jāyatī ṭhito hoti;}}\\
{\setstretch{.7}
\quad\EnglishColumn{no wave arises, but the ocean remains steady,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evaṃ ṭhito anejassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{so too he should be steady, without impulse;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ussadaṃ bhikkhu na kareyya kuhiñci”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{a bhikkhu should not cause a swelling anywhere.”}\
\par}
\end{samepage}

{\PaliColumn{927.}}\\

\begin{samepage}
{\PaliColumn{“akittayī vivaṭacakkhu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“The one with opened eyes declared}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sakkhidhammaṃ parissayavinayaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{the Dhamma he witnessed, the removal of obstacles.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paṭipadaṃ vadehi bhaddante,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Speak about the practice, venerable one,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pātimokkhaṃ athavāpi samādhiṃ”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the Pātimokkha and also concentration.”}\
\par}
\end{samepage}

{\PaliColumn{928.}}\\

\begin{samepage}
{\PaliColumn{“cakkhūhi neva lolassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One should not be restless with the eyes,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{gāmakathāya āvaraye sotaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should block the ears against village gossip.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rase ca nānugijjheyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should not be greedy for tastes,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na ca mamāyetha kiñci lokasmiṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{and one should not take anything in the world as ‘mine.’}\
\par}
\end{samepage}

{\PaliColumn{929.}}\\

\begin{samepage}
{\PaliColumn{“phassena yadā phuṭṭhassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“When he is touched by a contact,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paridevaṃ bhikkhu na kareyya kuhiñcñcci;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a bhikkhu should not lament anywhere.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhavañca nābhijappeyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He should not long for any state of existence,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bheravesu ca na sampavedheyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{nor should he tremble amid fearful conditions.}\
\par}
\end{samepage}

{\PaliColumn{930.}}\\

\begin{samepage}
{\PaliColumn{“annānamatho pānānaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having obtained food and drink,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{khādanīyānaṃ athopi vatthānaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{things to eat or clothes to wear,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{laddhā na sannidhiṃ kayirā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should not hoard them,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na ca parittase tāni alabhamāno.}}\\
{\setstretch{.7}
\quad\EnglishColumn{nor be agitated if he does not obtain them.}\
\par}
\end{samepage}

{\PaliColumn{931.}}\\

\begin{samepage}
{\PaliColumn{“jhāyī na pādalolassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He should be a meditator, not one with restless feet;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{virame kukkuccā nappamajjeyya;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should refrain from regret; he should not be heedless.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{athāsanesu sayanesu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{A bhikkhu should dwell amid seats and beds}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{appasaddesu bhikkhu vihareyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{[in lodgings] where there is little noise.}\
\par}
\end{samepage}

{\PaliColumn{932.}}\\

\begin{samepage}
{\PaliColumn{“niddaṃ na bahulīkareyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He should not overindulge in sleep;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{jāgariyaṃ bhajeyya ātāpī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should be devoted to wakefulness, ardent.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tandiṃ māyaṃ hassaṃ khiḍḍaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He should abandon indolence, hypocrisy, laughing, and playfulness,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{methunaṃ vippajahe savibhūsaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{sexual intercourse, and personal ornamentation.}\
\par}
\end{samepage}

{\PaliColumn{933.}}\\

\begin{samepage}
{\PaliColumn{“āthabbaṇaṃ supinaṃ lakkhaṇaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He should not employ Atharva spells,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{no vidahe athopi nakkhattaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{or interpret dreams, signs, or constellations.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{virutañca gabbhakaraṇaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{My follower would not decipher animal cries,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tikicchaṃ māmako na seveyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{or practice healing or making women fertile.}\
\par}
\end{samepage}

{\PaliColumn{934.}}\\

\begin{samepage}
{\PaliColumn{“nindāya nappavedheyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“A bhikkhu should not be shaken by blame,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na uṇṇameyya pasaṃsito bhikkhu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{nor should he be elated when praised.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{lobhaṃ saha macchariyena,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He should dispel greed and miserliness,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kodhaṃ pesuṇiyañca panudeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{as well as anger and slanderous speech.}\
\par}
\end{samepage}

{\PaliColumn{935.}}\\

\begin{samepage}
{\PaliColumn{“kayavikkaye na tiṭṭheyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“A bhikkhu should not engage in buying and selling;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{upavādaṃ bhikkhu na kareyya kuhiñci;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should not do anything that incurs criticism.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{gāme ca nābhisajjeyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He should not form intimacy in the village;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{lābhakamyā janaṃ na lapayeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should not address people from a desire for gain.}\
\par}
\end{samepage}

{\PaliColumn{936.}}\\

\begin{samepage}
{\PaliColumn{“na ca katthitā siyā bhikkhu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“A bhikkhu should not be boastful,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na ca vācaṃ payuttaṃ bhāseyya;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and he should not utter insinuating speech.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pāgabbhiyaṃ na sikkheyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He should not train himself in impudence;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kathaṃ viggāhikaṃ na kathayeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should not utter contentious talk.}\
\par}
\end{samepage}

{\PaliColumn{937.}}\\

\begin{samepage}
{\PaliColumn{“mosavajje na nīyetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He should not be led into false speech,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sampajāno saṭhāni na kayirā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{nor should he deliberately do anything deceptive.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atha jīvitena paññāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Because of his way of life, his wisdom,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sīlabbatena nāññamatimaññe.}}\\
{\setstretch{.7}
\quad\EnglishColumn{or his good behavior and observances, he should not despise another.}\
\par}
\end{samepage}

{\PaliColumn{938.}}\\

\begin{samepage}
{\PaliColumn{“sutvā rusito bahuṃ vācaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“When provoked, having heard many words}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{samaṇānaṃ vā puthujanānaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{from ascetics who speak profusely,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pharusena ne na paṭivajjā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should not respond to them harshly,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na hi santo paṭisenikaronti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for the good do not retaliate.}\
\par}
\end{samepage}

{\PaliColumn{939.}}\\

\begin{samepage}
{\PaliColumn{“etañca dhammamaññāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having understood this Dhamma, investigating,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vicinaṃ bhikkhu sadā sato sikkhe;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a bhikkhu should always train mindfully.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{santīti nibbutiṃ ñatvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having known quenching as peace,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sāsane gotamassa na pamajjeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should not be heedless in Gotama’s teaching.}\
\par}
\end{samepage}

{\PaliColumn{940.}}\\

\begin{samepage}
{\PaliColumn{“abhibhū hi so anabhibhūto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“For he is a conqueror who is himself unconquered;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sakkhidhammamanītihamadassī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he saw the Dhamma as a witness, not by hearsay.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tasmā hi tassa bhagavato sāsane,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Therefore, heedful in that Blessed One’s teaching,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{appamatto sadā namassamanusikkhe”ti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{always honoring it, one should train in accordance with it.”}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{tuvaṭakasuttaṃ cuddasamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{15. attadaṇḍasuttaṃ (KN 5.53)}}\\
{\setstretch{.7}
\quad\EnglishColumn{One Who has Taken up the Rod}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{941.}}\\

\begin{samepage}
{\PaliColumn{“attadaṇḍā bhayaṃ jātaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Fear has arisen from one who has taken up the rod:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{janaṃ passatha medhagaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{see the people engaged in strife.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saṃvegaṃ kittayissāmi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{I will tell you of my sense of urgency,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yathā saṃvijitaṃ mayā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{how I was stirred by a sense of urgency.}\
\par}
\end{samepage}

{\PaliColumn{942.}}\\

\begin{samepage}
{\PaliColumn{“phandamānaṃ pajaṃ disvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having seen the population trembling}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{macche appodake yathā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{like fish in a pool with little water,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aññamaññehi byāruddhe,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having seen them hostile to one another,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{disvā maṃ bhayamāvisi.}}\\
{\setstretch{.7}
\quad\EnglishColumn{fear came upon me.}\
\par}
\end{samepage}

{\PaliColumn{943.}}\\

\begin{samepage}
{\PaliColumn{“samantamasāro loko,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“The world was insubstantial all around;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{disā sabbā sameritā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{all the directions were in turmoil.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{icchaṃ bhavanamattano,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Desiring an abode for myself,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nāddasāsiṃ anositaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{I did not see [any place] unoccupied.}\
\par}
\end{samepage}

{\PaliColumn{944.}}\\

\begin{samepage}
{\PaliColumn{“osānetveva byāruddhe,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having seen those hostile at the end,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{disvā me aratī ahu;}}\\
{\setstretch{.7}
\quad\EnglishColumn{discontent came upon me.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{athettha sallamaddakkhiṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Then I saw the dart here,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{duddasaṃ hadayanissitaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{hard to see, nestled in the heart.}\
\par}
\end{samepage}

{\PaliColumn{945.}}\\

\begin{samepage}
{\PaliColumn{“yena sallena otiṇṇo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“When one is struck by that dart}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{disā sabbā vidhāvati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{one runs astray in all directions.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tameva sallamabbuyha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But having drawn out that dart,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na dhāvati na sīdati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one does not run, does not sink.”}\
\par}
\end{samepage}

{\PaliColumn{946.}}\\

\begin{samepage}
{\PaliColumn{“tattha sikkhānugīyanti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{There the trainings are recited:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yāni loke gadhitāni;}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Whatever bonds there are in the world,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na tesu pasuto siyā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not be intent on them.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nibbijjha sabbaso kāme.}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having entirely pierced through sensual pleasures,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sikkhe nibbānamattano.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should train for one’s own nibbāna.}\
\par}
\end{samepage}

{\PaliColumn{947.}}\\

\begin{samepage}
{\PaliColumn{“sacco siyā appagabbho,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One should be truthful, not impudent,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{amāyo rittapesuṇo;}}\\
{\setstretch{.7}
\quad\EnglishColumn{without hypocrisy, devoid of divisive speech.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{akkodhano lobhapāpaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Without anger, a muni should cross over}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vevicchaṃ vitare muni.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the evil of greed and miserliness.}\
\par}
\end{samepage}

{\PaliColumn{948.}}\\

\begin{samepage}
{\PaliColumn{“niddaṃ tandiṃ sahe thīnaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One should vanquish sleepiness, torpor, and dullness;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pamādena na saṃvase;}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not keep company with heedlessness.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atimāne na tiṭṭheyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{A person whose mind is set on nibbāna}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nibbānamanaso naro.}}\\
{\setstretch{.7}
\quad\EnglishColumn{should not persist in arrogance.}\
\par}
\end{samepage}

{\PaliColumn{949.}}\\

\begin{samepage}
{\PaliColumn{“mosavajje na nīyetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One should not be led into false speech;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rūpe snehaṃ na kubbaye;}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not engender affection for form.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mānañca parijāneyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should fully understand conceit,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sāhasā virato care.}}\\
{\setstretch{.7}
\quad\EnglishColumn{and one should refrain from rashness.}\
\par}
\end{samepage}

{\PaliColumn{950.}}\\

\begin{samepage}
{\PaliColumn{“purāṇaṃ nābhinandeyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One should not delight in the old;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nave khantiṃ na kubbaye;}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not form a liking for the new.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hiyyamāne na soceyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should not sorrow over what is diminishing;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ākāsaṃ na sito siyā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not be attached to an attraction.}\
\par}
\end{samepage}

{\PaliColumn{951.}}\\

\begin{samepage}
{\PaliColumn{“gedhaṃ brūmi mahoghoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“I call greed the great flood,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ājavaṃ brūmi jappanaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{I call longing the rapids,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ārammaṇaṃ pakappanaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{the basis, compulsion,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāmapaṅko duraccayo.}}\\
{\setstretch{.7}
\quad\EnglishColumn{the swamp of sensuality hard to overcome.}\
\par}
\end{samepage}

{\PaliColumn{952.}}\\

\begin{samepage}
{\PaliColumn{“saccā avokkamma muni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Not deviating from truth, a muni,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{thale tiṭṭhati brāhmaṇo;}}\\
{\setstretch{.7}
\quad\EnglishColumn{a brahmin, stands on high ground.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbaṃ so paṭinissajja,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having relinquished all,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa ve santoti vuccati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he is indeed called peaceful.}\
\par}
\end{samepage}

{\PaliColumn{953.}}\\

\begin{samepage}
{\PaliColumn{“sa ve vidvā sa vedagū,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One who truly is a knower, a Veda-master,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ñatvā dhammaṃ anissito;}}\\
{\setstretch{.7}
\quad\EnglishColumn{having known the Dhamma, is independent.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sammā so loke iriyāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Living rightly in the world,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na pihetīdha kassaci.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he does not long for anything here.}\
\par}
\end{samepage}

{\PaliColumn{954.}}\\

\begin{samepage}
{\PaliColumn{“yodha kāme accatari,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One here who has crossed over sensual pleasures,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saṅgaṃ loke duraccayaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{the tie so hard to overcome in the world,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na so socati nājjheti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{who has cut off the stream, without bonds,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{chinnasoto abandhano.}}\\
{\setstretch{.7}
\quad\EnglishColumn{does not sorrow, does not hope.}\
\par}
\end{samepage}

{\PaliColumn{955.}}\\

\begin{samepage}
{\PaliColumn{“yaṃ pubbe taṃ visosehi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Dry up what pertains to the past;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pacchā te māhu kiñcanaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{let there be nothing afterward.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{majjhe ce no gahessasi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{If you do not grasp in the middle,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{upasanto carissasi.}}\\
{\setstretch{.7}
\quad\EnglishColumn{you will live in peace.}\
\par}
\end{samepage}

{\PaliColumn{956.}}\\

\begin{samepage}
{\PaliColumn{“sabbaso nāmarūpasmiṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One who does not claim as ‘mine’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yassa natthi mamāyitaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{anything at all here in name-and-form,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{asatā ca na socati,}}\\
{\setstretch{.7}
\quad\EnglishColumn{who does not sorrow over what is nonexistent,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sa ve loke na jīyati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{truly does not lose out in the world.}\
\par}
\end{samepage}

{\PaliColumn{957.}}\\

\begin{samepage}
{\PaliColumn{“yassa natthi idaṃ meti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“One for whom nothing is taken}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{paresaṃ vāpi kiñcanaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{as ‘this is mine’ or ‘[this belongs] to others,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mamattaṃ so asaṃvindaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{not finding anything to be taken as ‘mine,’}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{natthi meti na socati.}}\\
{\setstretch{.7}
\quad\EnglishColumn{does not sorrow, thinking: ‘It is not mine.’}\
\par}
\end{samepage}

{\PaliColumn{958.}}\\

\begin{samepage}
{\PaliColumn{“aniṭṭhurī ananugiddho,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Not bitter, not greedy,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anejo sabbadhī samo;}}\\
{\setstretch{.7}
\quad\EnglishColumn{without impulse, everywhere the same —}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tamānisaṃsaṃ pabrūmi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{when asked about one who is unshakable,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pucchito avikampinaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{I call that the benefit.}\
\par}
\end{samepage}

{\PaliColumn{959.}}\\

\begin{samepage}
{\PaliColumn{“anejassa vijānato,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“For one without impulse, who understands,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{natthi kāci nisaṅkhati;}}\\
{\setstretch{.7}
\quad\EnglishColumn{there is no activation at all.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{virato so viyārabbhā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Abstaining from instigation,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{khemaṃ passati sabbadhi.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he sees security everywhere.}\
\par}
\end{samepage}

{\PaliColumn{960.}}\\

\begin{samepage}
{\PaliColumn{“na samesu na omesu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“The muni does not speak of himself}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na ussesu vadate muni;}}\\
{\setstretch{.7}
\quad\EnglishColumn{as among equals, inferiors, or superiors.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{santo so vītamaccharo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Peaceful, without miserliness,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nādeti na nirassatī”ti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he does not take, does not reject” —}\
\par}
\end{samepage}

{\EnglishColumn{so said the Blessed One.}}\\

{\PaliColumn{attadaṇḍasuttaṃ pannarasamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
\begin{samepage}
{\PaliColumn{16. sāriputtasuttaṃ (KN 5.54)}}\\
{\setstretch{.7}
\quad\EnglishColumn{Sāriputta}\
\par}
\end{samepage}

\vskip 0.1in
{\PaliColumn{961.}}\\

\begin{samepage}
{\PaliColumn{“na me diṭṭho ito pubbe,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“I have never before seen,”}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(iccāyasmā sāriputto)}}\\
{\setstretch{.7}
\quad\EnglishColumn{(said the Venerable Sāriputta),}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na suto uda kassaci;}}\\
{\setstretch{.7}
\quad\EnglishColumn{“nor have I ever heard it from anyone —}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{evaṃ vagguvado satthā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{a teacher of such lovely speech,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{tusitā gaṇimāgato.}}\\
{\setstretch{.7}
\quad\EnglishColumn{leader of a group, who has come from Tusita,}\
\par}
\end{samepage}

{\PaliColumn{962.}}\\

\begin{samepage}
{\PaliColumn{“sadevakassa lokassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“as the One with Vision is seen}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yathā dissati cakkhumā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{in the world together with its devas.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbaṃ tamaṃ vinodetvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having dispelled all darkness,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ekova ratimajjhagā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{alone, he achieved delight.}\
\par}
\end{samepage}

{\PaliColumn{963.}}\\

\begin{samepage}
{\PaliColumn{“taṃ buddhaṃ asitaṃ tādiṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“To the Buddha, unattached, impartial,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{akuhaṃ gaṇimāgataṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{not a schemer, one who has come as leader of a group,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bahūnamidha baddhānaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{I have come in need with a question}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atthi pañhena āgamaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{on behalf of the many here who are bound.}\
\par}
\end{samepage}

{\PaliColumn{964.}}\\

\begin{samepage}
{\PaliColumn{“bhikkhuno vijigucchato,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“When a bhikkhu is repulsed}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhajato rittamāsanaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{and resorts to a solitary seat,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rukkhamūlaṃ susānaṃ vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{to the foot of a tree or a charnel ground,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pabbatānaṃ guhāsu vā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{or [to a seat] in mountain caves,}\
\par}
\end{samepage}

{\PaliColumn{965.}}\\

\begin{samepage}
{\PaliColumn{“uccāvacesu sayanesu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“to various sorts of lodgings,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kīvanto tattha bheravā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{how many fearful things are there}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yehi bhikkhu na vedheyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{because of which a bhikkhu should not tremble}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nigghose sayanāsane.}}\\
{\setstretch{.7}
\quad\EnglishColumn{in a lodging with little noise?}\
\par}
\end{samepage}

{\PaliColumn{966.}}\\

\begin{samepage}
{\PaliColumn{“katī parissayā loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“How many obstacles in the world}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{gacchato agataṃ disaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{confront one going to the untraveled region,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ye bhikkhu abhisambhave,}}\\
{\setstretch{.7}
\quad\EnglishColumn{[obstacles] that a bhikkhu must overcome}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pantamhi sayanāsane.}}\\
{\setstretch{.7}
\quad\EnglishColumn{when living in a remote lodging?}\
\par}
\end{samepage}

{\PaliColumn{967.}}\\

\begin{samepage}
{\PaliColumn{“kyāssa byappathayo assu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“What ways of speech should he adopt?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kyāssassu idha gocarā;}}\\
{\setstretch{.7}
\quad\EnglishColumn{On what resorts here should he rely?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāni sīlabbatānāssu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{What should be the good behavior and observances}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pahitattassa bhikkhuno.}}\\
{\setstretch{.7}
\quad\EnglishColumn{of a bhikkhu who is resolute?}\
\par}
\end{samepage}

{\PaliColumn{968.}}\\

\begin{samepage}
{\PaliColumn{“kaṃ so sikkhaṃ samādāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having taken up what training,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ekodi nipako sato;}}\\
{\setstretch{.7}
\quad\EnglishColumn{dwelling unified, judicious, mindful,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kammāro rajatasseva,}}\\
{\setstretch{.7}
\quad\EnglishColumn{should he blow away his own stains}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{niddhame malamattano”.}}\\
{\setstretch{.7}
\quad\EnglishColumn{as a smith [removes the flaws] of silver?”}\
\par}
\end{samepage}

{\PaliColumn{969.}}\\

\begin{samepage}
{\PaliColumn{“vijigucchamānassa yadidaṃ phāsu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“As one who understands I will tell you}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{(sāriputtāti bhagavā)}}\\
{\setstretch{.7}
\quad\EnglishColumn{(Sāriputta,” said the Blessed One),}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rittāsanaṃ sayanaṃ sevato ce;}}\\
{\setstretch{.7}
\quad\EnglishColumn{“what is comfortable for one who is repulsed,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sambodhikāmassa yathānudhammaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{if, wishing for enlightenment, in accordance with the Dhamma,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{taṃ te pavakkhāmi yathā pajānaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he resorts to a solitary lodging.}\
\par}
\end{samepage}

{\PaliColumn{970.}}\\

\begin{samepage}
{\PaliColumn{“pañcannaṃ dhīro bhayānaṃ na bhāye,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“A wise one, a mindful bhikkhu of bounded conduct,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhikkhu sato sapariyantacārī;}}\\
{\setstretch{.7}
\quad\EnglishColumn{should not be afraid of five perils:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ḍaṃsādhipātānaṃ sarīsapānaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{gadflies and other flies, creeping serpents,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{manussaphassānaṃ catuppadānaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{encounters with people, four-footed animals,}\
\par}
\end{samepage}

{\PaliColumn{971.}}\\

\begin{samepage}
{\PaliColumn{“paradhammikānampi na santaseyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“nor should he fear followers of other teachings,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{disvāpi tesaṃ bahubheravāni;}}\\
{\setstretch{.7}
\quad\EnglishColumn{even when he sees the many fears they pose.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{athāparāni abhisambhaveyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One seeking the good should conquer}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{parissayāni kusalānuesī.}}\\
{\setstretch{.7}
\quad\EnglishColumn{any other obstacles [he encounters].}\
\par}
\end{samepage}

{\PaliColumn{972.}}\\

\begin{samepage}
{\PaliColumn{“ātaṅkaphassena khudāya phuṭṭho,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Afflicted by illness and hunger,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sītaṃ atuṇhaṃ adhivāsayeyya;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should endure cold and excessive heat.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{so tehi phuṭṭho bahudhā anoko,}}\\
{\setstretch{.7}
\quad\EnglishColumn{The homeless one, affected by these in many ways,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vīriyaṃ parakkammadaḷhaṃ kareyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{should remain firm, exerting energy.}\
\par}
\end{samepage}

{\PaliColumn{973.}}\\

\begin{samepage}
{\PaliColumn{“theyyaṃ na kāre na musā bhaṇeyya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He should not commit theft or speak falsely;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mettāya phasse tasathāvarāni;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should suffuse the frail and firm with loving-kindness.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yadāvilattaṃ manaso vijaññā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{If he should recognize any blemish of the mind,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kaṇhassa pakkhoti vinodayeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should dispel it as ‘a faction of the Dark One.’}\
\par}
\end{samepage}

{\PaliColumn{974.}}\\

\begin{samepage}
{\PaliColumn{“kodhātimānassa vasaṃ na gacche,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“He should not come under the control of anger and arrogance,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mūlampi tesaṃ palikhañña tiṭṭhe;}}\\
{\setstretch{.7}
\quad\EnglishColumn{but should dig them up by their root too.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{athappiyaṃ vā pana appiyaṃ vā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Then, withstanding the pleasing and displeasing,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{addhā bhavanto abhisambhaveyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should conquer them as well.}\
\par}
\end{samepage}

{\PaliColumn{975.}}\\

\begin{samepage}
{\PaliColumn{“paññaṃ purakkhatvā kalyāṇapīti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Giving precedence to wisdom, rejoicing in the good,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vikkhambhaye tāni parissayāni;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should suppress those obstacles.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aratiṃ sahetha sayanamhi pante,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He should vanquish discontent with remote lodgings;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{caturo sahetha paridevadhamme.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should vanquish the four factors of lamentation.}\
\par}
\end{samepage}

{\PaliColumn{976.}}\\

\begin{samepage}
{\PaliColumn{“kiṃsū asissāmi kuvaṃ vā asissaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“ ‘What will I eat? Where will I eat?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dukkhaṃ vata settha kvajja sessaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{Alas, I slept badly! Where will I sleep?’:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ete vitakke paridevaneyye,}}\\
{\setstretch{.7}
\quad\EnglishColumn{one in training, wandering without abode,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vinayetha sekho aniketacārī.}}\\
{\setstretch{.7}
\quad\EnglishColumn{should remove these thoughts of lamentation.}\
\par}
\end{samepage}

{\PaliColumn{977.}}\\

\begin{samepage}
{\PaliColumn{“annañca laddhā vasanañca kāle,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Having obtained a meal and clothing in time,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mattaṃ so jaññā idha tosanatthaṃ;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should know moderation for the sake of contentment here.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{so tesu gutto yatacāri gāme,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Guarded in these things, wandering restrained in the village,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rusitopi vācaṃ pharusaṃ na vajjā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{even when provoked, he should not utter harsh speech.}\
\par}
\end{samepage}

{\PaliColumn{978.}}\\

\begin{samepage}
{\PaliColumn{“okkhittacakkhu na ca pādalolo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“His eyes should be downcast; he should not have restless feet;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{jhānānuyutto bahujāgarass;}}\\
{\setstretch{.7}
\quad\EnglishColumn{intent on jhāna, he should be wakeful.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{upekkhamārabbha samāhitatto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Inwardly concentrated, based on equanimity,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{takkāsayaṃ kukkucciyūpachinde.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should cut off regret and inclination to thought.}\
\par}
\end{samepage}

{\PaliColumn{979.}}\\

\begin{samepage}
{\PaliColumn{“cudito vacībhi satimābhinande,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“If reproached with words, a mindful one should welcome them;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabrahmacārīsu khilaṃ pabhinde;}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should break through a barren heart toward fellow monastics.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vācaṃ pamuñce kusalaṃ nātivelaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{He should utter wholesome speech, but not to excess;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{janavādadhammāya na cetayeyya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{he should not intend [speech] that people might blame.}\
\par}
\end{samepage}

{\PaliColumn{980.}}\\

\begin{samepage}
{\PaliColumn{“athāparaṃ pañca rajāni loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“Further, there are five kinds of dust in the world,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yesaṃ satīmā vinayāya sikkhe;}}\\
{\setstretch{.7}
\quad\EnglishColumn{for the removal of which a mindful one should train.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rūpesu saddesu atho rasesu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should vanquish lust for forms and sounds,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{gandhesu phassesu sahetha rāgaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{for tastes, odors, and tactile objects.}\
\par}
\end{samepage}

{\PaliColumn{981.}}\\

\begin{samepage}
{\PaliColumn{“etesu dhammesu vineyya chandaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“A bhikkhu who is mindful, well liberated in mind,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{bhikkhu satimā suvimuttacitto;}}\\
{\setstretch{.7}
\quad\EnglishColumn{should remove desire for these things.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kālena so sammā dhammaṃ parivīmaṃsamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{At the proper time rightly investigating the Dhamma,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ekodibhūto vihane tamaṃ so”ti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{unified, he should destroy darkness” —}\
\par}
\end{samepage}

{\EnglishColumn{so said the Blessed One.}}\\

{\PaliColumn{sāriputtasuttaṃ soḷasamaṃ niṭṭhitaṃ.}}\\

\vskip 0.1in
{\PaliColumn{aṭṭhakavaggo catuttho niṭṭhito.}}\\

{\PaliColumn{tassuddānaṃ —}}\\

{\PaliColumn{kāmaṃ guhañca duṭṭhā ca,}}\\

{\PaliColumn{suddhañca paramā jarā.}}\\

{\PaliColumn{metteyyo ca pasūro ca,}}\\

{\PaliColumn{māgaṇḍi purābhedanaṃ.}}\\

{\PaliColumn{kalahaṃ dve ca byūhāni,}}\\

{\PaliColumn{punadeva tuvaṭṭakaṃ.}}\\

{\PaliColumn{attadaṇḍavaraṃ suttaṃ,}}\\

{\PaliColumn{therapuṭṭhena soḷasa.}}\\

{\PaliColumn{iti etāni suttāni,}}\\

{\PaliColumn{sabbānaṭṭhakavaggikāti.}}\\
