@ eng-an-intro.tex
Venerables, these two uncertain cases come up for recitation.

@ eng-an-01.tex
If any bhikkhu should sit down together with a woman, one (man) with one (woman), privately, on a concealed seat (that is) sufficiently fit for doing (it), (and then if) a female lay-follower whose words can be trusted having seen that, should speak according to one of three cases; according to disqualification, according to what concerns the community in the beginning and in the rest (of the procedure), or according to expiation,

@ eng-an-01a.tex
(then) the bhikkhu who is admitting the sitting down should be made to do (what is) according to one of three cases; according to disqualification, or according to what concerns the community in the beginning and in the rest (of the procedure), or according to expiation, or according to whatever that female lay-follower whose words can be trusted should say, according to that the bhikkhu is to be made to do. This is an uncertain case.

@ eng-an-02.tex
But even if the seat is neither concealed nor sufficiently fit for doing it, but is sufficient for speaking suggestively to a woman with depraved words:

@ eng-an-02a.tex
if any bhikkhu should sit down on such a seat together with a woman—one (man) with one (woman), privately—(and then if) a female lay-follower whose words can be trusted having seen that, should speak according to one of two cases; according to what concerns the community in the beginning and in the rest, or according to expiation,

@ eng-an-02b.tex
(then) the bhikkhu admitting the sitting down is to be made to do according to one of two cases; according to what concerns the community in the beginning and in the rest (of the procedure), or according to expiation, or according to whatever that female lay-follower whose words can be trusted should say, according to that the bhikkhu is to be made to do, this too is an uncertain case.

@ eng-an-uddittha.tex
Venerables, the two uncertain cases have been recited. +

@ eng-an-parisuddha.tex
Concerning that I ask the Venerables: (Are you) pure in this? +
A second time again I ask: (Are you) pure in this? +
A third time again I ask: (Are you) pure in this? +
The venerables are pure in this, therefore there is silence, thus I bear this (in mind).

@ eng-center-an-nitthito.tex
The recitation of the uncertain (cases) is finished.
