
@ eng-center-pk-0.tex
Uposatha Preliminary Duties

@ eng-pk-nava.tex
Before doing the Uposatha the ninefold duty should be done:

@ eng-pk-01.tex
the sweeping of that place;

@ eng-pk-02.tex
the lighting of the lamp there;

@ eng-pk-03.tex
the spreading of the sitting mats;

@ eng-pk-04.tex
preparing water for drinking and using;

@ eng-pk-05.tex
the bringing of the consent of the bhikkhus who are deserving (to give their) consent;

@ eng-pk-06.tex
and of those bhikkhus who do not participate in the Uposatha the bringing of purity too;

@ eng-pk-07.tex
the telling of the season;

@ eng-pk-08.tex
the counting of the bhikkhus,

@ eng-pk-09.tex
and the instructing of the bhikkhunīs.

@ eng-pk-tattha.tex
Herein with regards the first four duties there is no duty of (lighting) the lamp because of the state of there being sunlight now, the three other ones have been done and completed by bhikkhus who know the bhikkhus' protocol.

@ eng-pk-aharana.tex
There is no bringing of consent (and) bringing of purity because of the absence in this boundary of seated bhikkhus who have left arm's length.

@ eng-pk-utukkhanam.tex
The so-called “telling the season” is declaring the season thus: this number (of Uposathas) have passed, this number (of Uposathas) are left.

@ eng-pk-utunidha.tex
In this dispensation there are three seasons, (namely) winter, summer,and rains. This is the winter season, and in this season there are eight Uposathas. With this fortnight one Uposatha has arrived, two Uposathas have past, five Uposathas are left. Thus the telling of the season is to be borne in mind by all the venerables.

@ eng-pk-evambhante.tex
(Yes, venerable sir)

@ eng-pk-ganana.tex
The so-called “counting of the bhikkhus” is the counting of the bhikkhus (thus): “In this Uposatha-hall so many bhikkhus have gathered for the purpose of (performing) the Uposatha.” 
In this Uposatha-hall four bhikkhus have gathered.

@ eng-pk-evambhante.tex
(Yes, venerable sir)

@ eng-pk-ovado.tex
There is no instruction of the bhikkhunīs now because of their non-existence. 

@ eng-pk-iti.tex
Thus, because of the state of having been done of the preliminary duties, which had an opportunity of doing (them), and because of the naturalness of the preliminary duties which had no opportunity of doing (them), the nine-fold Preliminary Function has been completed.

@ eng-pk-nitthite.tex
When the preliminary duty has been finished:

@ eng-pk-nitthite0.tex
If the day is a certain one of the fourteenth (day Uposathas), fifteenth (day Uposathas), or the Unity (-Uposathas), like today is the fifteenth day Uposatha.

@ eng-pk-11.tex
Whatever bhikkhus are entitled (to carry out the legal) act, who are qualified for the Uposatha of the Community, four or more than that, regular, who have not committed an offence of pārājika, who have not been suspended by the Community,

@ eng-pk-12.tex
and they, not having left arms-length, remain in one boundary,

@ eng-pk-13.tex
and if among them common offences, founded on eating at the wrong time and so on, are not found,

@ eng-pk-14.tex
and if among them within arms-length there is no person fit to be excluded, who is to be excluded from arms-length by making (him go) outside,

@ eng-pk-evantam.tex
then that (legal) act of Uposatha endowed with these four characteristics is called “(one that) has reached suitability,” (and is) fit to perform.

@ eng-pk-uposathakammassa.tex
Having known the state of reached suitability of the (legal) act of Uposatha, it is to be approved of by the Community (that) the Uposatha is being done now.”

@ eng-pk-evambhante.tex
(Yes, venerable sir)

@ eng-pk-senior.tex
Elder: “Having made to undertake the preparations and preliminary duties I make the invitation to recite the Disciplinary Code with the approval of the seated community.”
