
\begin{samepage}
{\PaliColumn{Khuddaka Nikāya, suttanipātapāḷi, 1. uragavaggo n, 3. khaggavisāṇasuttaṃ (KN 5.3)}}\\
{\setstretch{.7}
\quad\EnglishColumn{3 THE RHINOCEROS HORN (KHAGGAVISĀṆA SUTTA)}\
\par}
\end{samepage}

{\PaliColumn{35.}}\\

\begin{samepage}
{\PaliColumn{sabbesu bhūtesu nidhāya daṇḍaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having put down the rod toward all beings,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aviheṭhayaṃ aññatarampi tesaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not harming a single one among them,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{na puttamiccheyya kuto sahāyaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should not desire a son, how then a companion?}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{eko care khaggavisāṇakappo.}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should live alone like a rhinoceros horn.}\
\par}
\end{samepage}

{\PaliColumn{36.}}\\

\begin{samepage}
{\PaliColumn{saṃsaggajātassa bhavanti snehā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{For one who has formed bonds, there is affection,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{snehanvayaṃ dukkhamidaṃ pahoti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{following on affection, this suffering arises.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ādīnavaṃ snehajaṃ pekkhamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Discerning the danger born of affection,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{37.}}\\

\begin{samepage}
{\PaliColumn{mitte suhajje anukampamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Sympathizing with friends dear to one’s heart,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hāpeti atthaṃ paṭibaddhacitto.}}\\
{\setstretch{.7}
\quad\EnglishColumn{with mind attached, one forsakes the good.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etaṃ bhayaṃ santhave pekkhamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Seeing this peril in intimacy,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{38.}}\\

\begin{samepage}
{\PaliColumn{vaṃso visālova yathā visatto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{As widespread bamboo becomes entwined,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{puttesu dāresu ca yā apekkhā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{just so is concern for wives and sons.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vaṃsakkaḷīrova sajjamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But like a bamboo shoot, not getting stuck,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{39.}}\\

\begin{samepage}
{\PaliColumn{migo araññamhi yathā abaddho,}}\\
{\setstretch{.7}
\quad\EnglishColumn{As a deer unbound in the forest}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yenicchakaṃ gacchati gocarāya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{goes off to graze wherever it wants,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{viññū naro seritaṃ pekkhamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{so a wise person, looking out for freedom,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{40.}}\\

\begin{samepage}
{\PaliColumn{āmantanā hoti sahāyamajjhe,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One is addressed in the midst of companions,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vāse ṭhāne gamane cārikāya.}}\\
{\setstretch{.7}
\quad\EnglishColumn{whether resting, standing, going, or traveling.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anabhijjhitaṃ seritaṃ pekkhamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Looking out for the freedom that is not coveted,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{41.}}\\

\begin{samepage}
{\PaliColumn{khiḍḍā ratī hoti sahāyamajjhe,}}\\
{\setstretch{.7}
\quad\EnglishColumn{There is play and delight in the midst of companions,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{puttesu ca vipulaṃ hoti pemaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{and affection for one’s sons is vast.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{piyavippayogaṃ vijigucchamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Averse to separation from those who are dear,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{42.}}\\

\begin{samepage}
{\PaliColumn{cātuddiso appaṭigho ca hoti,}}\\
{\setstretch{.7}
\quad\EnglishColumn{At home in the four directions, unrepelled,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{santussamāno itarītarena.}}\\
{\setstretch{.7}
\quad\EnglishColumn{contented with anything whatsoever,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{parissayānaṃ sahitā achambhī,}}\\
{\setstretch{.7}
\quad\EnglishColumn{enduring obstacles, fearless,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{43.}}\\

\begin{samepage}
{\PaliColumn{dussaṅgahā pabbajitāpi eke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Even some monastics are hard to please;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{atho gahaṭṭhā gharamāvasantā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{so, too, householders living at home.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{appossukko paraputtesu hutvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Being unconcerned about others’ sons,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{44.}}\\

\begin{samepage}
{\PaliColumn{oropayitvā gihibyañjanāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having discarded the marks of a layman}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sañchinnapatto yathā koviḷāro.}}\\
{\setstretch{.7}
\quad\EnglishColumn{like a koviḷāra tree whose leaves are shed,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{chetvāna vīro gihibandhanāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having cut off a layman’s bonds, the hero}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{45.}}\\

\begin{samepage}
{\PaliColumn{sace labhetha nipakaṃ sahāyaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{If one should find a judicious companion,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saddhiṃ caraṃ sādhuvihāridhīraṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{a fellow wanderer, of good behavior, resolute,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{abhibhuyya sabbāni parissayāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having overcome all obstacles, one should}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{careyya tenattamano satīmā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{live with him, satisfied and mindful.}\
\par}
\end{samepage}

{\PaliColumn{46.}}\\

\begin{samepage}
{\PaliColumn{no ce labhetha nipakaṃ sahāyaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{But if one does not find a judicious companion,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saddhiṃ caraṃ sādhuvihāridhīraṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{a fellow wanderer, of good behavior, resolute,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rājāva raṭṭhaṃ vijitaṃ pahāya,}}\\
{\setstretch{.7}
\quad\EnglishColumn{like a king who has abandoned a conquered realm,}\
\par}
\end{samepage}

{\PaliColumn{eko care mātaṅgaraññeva nāgo.}}\\

{\PaliColumn{47.}}\\

\begin{samepage}
{\PaliColumn{addhā pasaṃsāma sahāyasampadaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Surely, we praise the excellence of companionship:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{seṭṭhā samā sevitabbā sahāyā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should resort to companions one’s equal or better.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ete aladdhā anavajjabhojī,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not obtaining these, as one who eats blamelessly}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{48.}}\\

\begin{samepage}
{\PaliColumn{disvā suvaṇṇassa pabhassarāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having seen radiant [bracelets] of gold,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kammāraputtena suniṭṭhitāni.}}\\
{\setstretch{.7}
\quad\EnglishColumn{skillfully fashioned by a goldsmith,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saṅghaṭṭamānāni duve bhujasmiṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{clashing together in pairs on the arm,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{49.}}\\

\begin{samepage}
{\PaliColumn{evaṃ dutiyena sahā mamassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Thus if I had a partner, I would incur}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vācābhilāpo abhisajjanā vā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{[fond] words of address or verbal friction.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etaṃ bhayaṃ āyatiṃ pekkhamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Looking out for this peril in the future,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{50.}}\\

\begin{samepage}
{\PaliColumn{kāmā hi citrā madhurā manoramā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Sensual pleasures are colorful, sweet, delightful,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{virūparūpena mathenti cittaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{but in their diversity they agitate the mind.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ādīnavaṃ kāmaguṇesu disvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having seen danger in the strands of sensual pleasure,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{51.}}\\

\begin{samepage}
{\PaliColumn{ītī ca gaṇḍo ca upaddavo ca,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“This is adversity, a boil, disaster,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rogo ca sallañca bhayañca metaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{an illness, a dart, and peril for me”:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{etaṃ bhayaṃ kāmaguṇesu disvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having seen this peril in the strands of sensual pleasure,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{52.}}\\

\begin{samepage}
{\PaliColumn{sītañca uṇhañca khudaṃ pipāsaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Cold and heat, hunger, thirst,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vātātape ḍaṃsasarīsape ca.}}\\
{\setstretch{.7}
\quad\EnglishColumn{wind, the hot sun, gadflies, and serpents:}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbānipetāni abhisambhavitvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having patiently endured all these,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{53.}}\\

\begin{samepage}
{\PaliColumn{nāgova yūthāni vivajjayitvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{As an elephant that has abandoned the herd —}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sañjātakhandho padumī uḷāro.}}\\
{\setstretch{.7}
\quad\EnglishColumn{with massive back, lotus-like, eminent —}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yathābhirantaṃ viharaṃ araññe,}}\\
{\setstretch{.7}
\quad\EnglishColumn{may live in the forest as he pleases,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{54.}}\\

\begin{samepage}
{\PaliColumn{aṭṭhānataṃ saṅgaṇikāratassa,}}\\
{\setstretch{.7}
\quad\EnglishColumn{It is impossible that one who delights in company}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{yaṃ phassaye sāmayikaṃ vimuttiṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{might attain even temporary liberation.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ādiccabandhussa vaco nisamma,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having attended to the word of the Kinsman of the Sun,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{55.}}\\

\begin{samepage}
{\PaliColumn{diṭṭhīvisūkāni upātivatto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“I have transcended the contortions of views,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{patto niyāmaṃ paṭiladdhamaggo.}}\\
{\setstretch{.7}
\quad\EnglishColumn{reached the fixed course, obtained the path.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{uppannañāṇomhi anaññaneyyo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{I have aroused knowledge, I’m not to be led by others”:}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{56.}}\\

\begin{samepage}
{\PaliColumn{nillolupo nikkuho nippipāso,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Without greed, without scheming, without thirst,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nimmakkho niddhantakasāvamoho.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not denigrating, with stains and delusion blown away,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nirāsayo sabbaloke bhavitvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{without wishes for anything in all the world,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{57.}}\\

\begin{samepage}
{\PaliColumn{pāpaṃ sahāyaṃ parivajjayetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should avoid an evil companion,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anatthadassiṃ visame niviṭṭhaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{who shows what is harmful, one settled in unrighteousness.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sayaṃ na seve pasutaṃ pamattaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should not freely associate with one who is intent and heedless;}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{58.}}\\

\begin{samepage}
{\PaliColumn{bahussutaṃ dhammadharaṃ bhajetha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{One should resort to the learned, a bearer of Dhamma,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{mittaṃ uḷāraṃ paṭibhānavantaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{an eminent friend gifted with ingenuity.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aññāya atthāni vineyya kaṅkhaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having known the benefits and removed doubt,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{59.}}\\

\begin{samepage}
{\PaliColumn{khiḍḍaṃ ratiṃ kāmasukhañca loke,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having found no satisfaction in the world}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{analaṅkaritvā anapekkhamāno.}}\\
{\setstretch{.7}
\quad\EnglishColumn{with play, delight, and sensual pleasures, not taking any interest in them,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vibhūsanaṭṭhānā virato saccavādī,}}\\
{\setstretch{.7}
\quad\EnglishColumn{refraining from ornaments, a speaker of truth,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{60.}}\\

\begin{samepage}
{\PaliColumn{puttañca dāraṃ pitarañca mātaraṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having abandoned children and wife,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dhanāni dhaññāni ca bandhavāni.}}\\
{\setstretch{.7}
\quad\EnglishColumn{father and mother, wealth, grain, and relatives,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{hitvāna kāmāni yathodhikāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{sensual pleasures according to the limit,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{61.}}\\

\begin{samepage}
{\PaliColumn{saṅgo eso parittamettha sokhyaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{“This is a tie, the happiness here is slight,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{appassādo dukkhamettha bhiyyo.}}\\
{\setstretch{.7}
\quad\EnglishColumn{giving little gratification; the suffering here is more,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{gaḷo eso iti ñatvā mutīmā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{this is a hook”: having known thus, a thoughtful person}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{62.}}\\

\begin{samepage}
{\PaliColumn{sandālayitvāna saṃyojanāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having sundered the fetters,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{jālaṃva bhetvā salilambucārī.}}\\
{\setstretch{.7}
\quad\EnglishColumn{like a fish in the water that has broken a net,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aggīva daḍḍhaṃ anivattamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{like a fire not returning to what has been burnt,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{63.}}\\

\begin{samepage}
{\PaliColumn{okkhittacakkhū na ca pādalolo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{With downcast gaze, not footloose,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{guttindriyo rakkhitamānasāno.}}\\
{\setstretch{.7}
\quad\EnglishColumn{with guarded faculties, with protected mind,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anavassuto apariḍayhamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{unpolluted, not feverish with passion,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{64.}}\\

\begin{samepage}
{\PaliColumn{ohārayitvā gihibyañjanāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having cast off the marks of a layman}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sañchannapatto yathā pārichatto.}}\\
{\setstretch{.7}
\quad\EnglishColumn{like a pāricchattaka tree that has shed its leaves,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kāsāyavattho abhinikkhamitvā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{clothed in ochre robes, having renounced,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{65.}}\\

\begin{samepage}
{\PaliColumn{rasesu gedhaṃ akaraṃ alolo,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not arousing greed for tastes, not hankering for them;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anaññaposī sapadānacārī.}}\\
{\setstretch{.7}
\quad\EnglishColumn{not nourishing others, walking for alms without skipping houses;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{kule kule appaṭibaddhacitto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{with a mind unbound to this or that family,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{66.}}\\

\begin{samepage}
{\PaliColumn{pahāya pañcāvaraṇāni cetaso,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having abandoned the five obstructions of mind,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{upakkilese byapanujja sabbe.}}\\
{\setstretch{.7}
\quad\EnglishColumn{having dispelled all mental defilements,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{anissito chetva sinehadosaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{independent, having cut off affection and hatred,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{67.}}\\

\begin{samepage}
{\PaliColumn{vipiṭṭhikatvāna sukhaṃ dukhañca,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having left behind pleasure and pain}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{pubbeva ca somanassadomanassaṃ.}}\\
{\setstretch{.7}
\quad\EnglishColumn{and previously [discarded] joy and dejection,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{laddhānupekkhaṃ samathaṃ visuddhaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having gained purified equanimity and serenity,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{68.}}\\

\begin{samepage}
{\PaliColumn{āraddhavīriyo paramatthapattiyā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{With energy aroused to attain the supreme goal,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{alīnacitto akusītavutti.}}\\
{\setstretch{.7}
\quad\EnglishColumn{with unsluggish mind and robust practice,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{daḷhanikkamo thāmabalūpapanno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{firmly persistent, equipped with strength and power,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{69.}}\\

\begin{samepage}
{\PaliColumn{paṭisallānaṃ jhānamariñcamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Not neglecting seclusion and jhāna,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{dhammesu niccaṃ anudhammacārī.}}\\
{\setstretch{.7}
\quad\EnglishColumn{always acting in accordance with the teachings,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{ādīnavaṃ sammasitā bhavesu,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having explored the danger in states of existence,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{70.}}\\

\begin{samepage}
{\PaliColumn{taṇhakkhayaṃ patthayamappamatto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Yearning for craving’s destruction, heedful,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{aneḷamūgo sutavā satīmā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{intelligent, learned, mindful,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{saṅkhātadhammo niyato padhānavā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{having comprehended the Dhamma, fixed in destiny, vigorous in striving,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{71.}}\\

\begin{samepage}
{\PaliColumn{sīhova saddesu asantasanto,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Like a lion unalarmed among sounds,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{vātova jālamhi asajjamāno.}}\\
{\setstretch{.7}
\quad\EnglishColumn{like the wind not caught in a net,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{padumaṃva toyena alippamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{untainted like a lotus by water,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{72.}}\\

\begin{samepage}
{\PaliColumn{sīho yathā dāṭhabalī pasayha,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Like the lion, king of beasts, who has fangs as its strength,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{rājā migānaṃ abhibhuyya cārī.}}\\
{\setstretch{.7}
\quad\EnglishColumn{who lives by attacking and overpowering,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sevetha pantāni senāsanāni,}}\\
{\setstretch{.7}
\quad\EnglishColumn{one should resort to remote lodgings;}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{73.}}\\

\begin{samepage}
{\PaliColumn{mettaṃ upekkhaṃ karuṇaṃ vimuttiṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{At the right time pursuing liberation by loving-kindness,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{āsevamāno muditañca kāle.}}\\
{\setstretch{.7}
\quad\EnglishColumn{equanimity, compassion, and altruistic joy,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sabbena lokena avirujjhamāno,}}\\
{\setstretch{.7}
\quad\EnglishColumn{not antagonized by the whole world,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{74.}}\\

\begin{samepage}
{\PaliColumn{rāgañca dosañca pahāya mohaṃ,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Having abandoned lust, hatred, and delusion,}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{sandālayitvāna saṃyojanāni.}}\\
{\setstretch{.7}
\quad\EnglishColumn{having sundered the fetters [that keep one bound],}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{asantasaṃ jīvitasaṅkhayamhi,}}\\
{\setstretch{.7}
\quad\EnglishColumn{not terrified at the extinction of life,}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\

{\PaliColumn{75.}}\\

\begin{samepage}
{\PaliColumn{bhajanti sevanti ca kāraṇatthā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{They resort to you and serve you for a motive;}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{nikkāraṇā dullabhā ajja mittā.}}\\
{\setstretch{.7}
\quad\EnglishColumn{friends without motive are today very rare.}\
\par}
\end{samepage}

\begin{samepage}
{\PaliColumn{attaṭṭhapaññā asucī manussā,}}\\
{\setstretch{.7}
\quad\EnglishColumn{Impure people are wise about their own good:}\
\par}
\end{samepage}

{\PaliColumn{eko care khaggavisāṇakappo.}}\\
