@ eng-sk-0.tex
Venerables, these cases related to the training come up for recitation.

@ eng-sk-1.tex
1. I shall wear (the under-robe) even all around,” thus the training is to be done.\\
2. I shall wrap (the outer-robes) even all around,” thus the training is to be done.

@ eng-sk-3.tex
3. I shall go well covered inside an inhabited area,” thus the training is to be done.\\
4. I shall sit well covered inside an inhabited area,” thus the training is to be done.

@ eng-sk-5.tex
5. I shall go well-restrained inside an inhabited area,” thus the training is to be done.\\
6. I shall sit well-restrained inside an inhabited area,” thus the training is to be done.

@ eng-sk-7.tex
7. I shall go with the eyes cast down inside an inhabited area,” thus the training is to be done.\\
8. I shall sit with the eyes cast down inside an inhabited area,” thus the training is to be done.

@ eng-sk-9.tex
9. I shall not go with (robes) lifted up inside an inhabited area,” thus the training is to be done.\\
10. I shall not sit with (robes) lifted up inside an inhabited area,” thus the training is to be done.

@ eng-sk-11.tex
11. I shall not go with loud laughter inside an inhabited area,” thus the training is to be done.\\
12. I shall not sit with loud laughter inside an inhabited area,” thus the training is to be done.

@ eng-sk-13.tex
13. I shall go quiet(ly) inside an inhabited area,” thus the training is to be done.\\
14. I shall sit quiet(ly) inside an inhabited area,” thus the training is to be done.

@ eng-sk-15.tex
15. I shall not go swaying the body inside an inhabited area,” thus the training is to be done.\\
16. I shall not sit swaying the body inside an inhabited area,” thus the training is to be done.

@ eng-sk-17.tex
17. I shall not go swaying the arms inside an inhabited area,” thus the training is to be done.\\
18. I shall not sit swaying the arms inside an inhabited area,” thus the training is to be done.

@ eng-sk-19.tex
19. I shall not go swaying the head inside an inhabited area,” thus the training is to be done.\\
20. I shall not sit swaying the head inside an inhabited area,” thus the training is to be done.

@ eng-sk-21.tex
21. I shall not go having made (the arms) a prop inside an inhabited area,” thus the training is to be done.\\
22. I shall not sit having made (the arms) a prop inside an inhabited area,” thus the training is to be done.

@ eng-sk-23.tex
23. I shall not go with (the head) covered inside an inhabited area,” thus the training is to be done.\\
24. I shall not sit with (the head) covered inside an inhabited area,” thus the training is to be done.

@ eng-sk-25.tex
25. I shall not go in a crouching (posture) inside an inhabited area,” thus the training is to be done.

@ eng-sk-26.tex
26. I shall not sit with the (knees) clasped-around inside an inhabited area,” thus the training is to be done.

@ eng-sk-chabbisati.tex
(Here ends) the Twenty-Six on Proper Behavior

@ eng-sk-27.tex
27. I shall accept alms-food appreciatively,” thus the training is to be done.

@ eng-sk-28.tex
28. I shall accept alms-food paying attention to the bowl,” thus the training is to be done.

@ eng-sk-29.tex
29. I shall accept alms-food which has curry in the proper proportion,” thus the training is to be done.

@ eng-sk-30.tex
30. I shall accept alms-food which is level with the rim,” thus the training is to be done.

@ eng-sk-31.tex
31. I shall eat alms-food appreciatively,” thus the training is to be done.

@ eng-sk-32.tex
32. I shall eat alms-food paying attention to the bowl,” thus the training is to be done.

@ eng-sk-33.tex
33. I shall eat alms-food systematically,” thus the training is to be done.

@ eng-sk-34.tex
34. I shall eat alms-food which has curry in the proper proportion,” thus the training is to be done.

@ eng-sk-35.tex
35. I shall not eat alms-food, having pressed (it) down into a shall heap,” thus the training is to be done.

@ eng-sk-36.tex
36. I shall not cover curry or condiment with rice out of liking for more,” thus the training is to be done.

@ eng-sk-37.tex
37. I shall not eat curry or rice, (when) not ill , having requested (it) for his own benefit, thus the training is to be done.

@ eng-sk-38.tex
38. I shall not look at another's bowl finding fault,” thus the training is to be done.

@  eng-sk-39.tex
39. I shall not make an over-large morsel (of food),” thus the training is to be done.

@  eng-sk-40.tex
40. I shall eat a round piece (of food),” thus the training is to be done.

@  eng-sk-41.tex
41. I shall not open the mouth when the morsel (of food) has not been brought to (it),” thus the training is to be done.

@  eng-sk-42.tex
42. I shall not put the whole hand onto the mouth while eating ,” thus the training is to be done.

@  eng-sk-43.tex
43. I shall not speak with a mouth which has a morsel (of food in it),” , thus the training is to be done.

@  eng-sk-44.tex
44. I shall not eat tossing up bits (of food),” thus the training is to be done.

@  eng-sk-45.tex
45. I shall not eat biting off a morsel (of food),” thus the training is to be done.

@  eng-sk-46.tex
46. I shall not eat puffing up (the cheeks),” thus the training is to be done.

@  eng-sk-47.tex
47. I shall not eat shaking (food) off the hand,” thus the training is to be done.

@  eng-sk-48.tex
48. I shall not eat scattering rice-grains,” thus the training is to be done.

@  eng-sk-49.tex
49. I shall not eat sticking out the tongue,” thus the training is to be done.

@  eng-sk-50.tex
50. I shall not eat making chomping (sounds),” thus the training is to be done.

@  eng-sk-51.tex
51. I shall not eat making slurping (sounds),” thus the training is to be done.

@  eng-sk-52.tex
52. I shall not eat licking the hand,” thus the training is to be done.

@  eng-sk-53.tex
53. I shall not eat licking the bowl,” thus the training is to be done.

@  eng-sk-54.tex
54. I shall not eat licking the lip(s),” thus the training is to be done.

@  eng-sk-55.tex
55. I shall not accept a drinking-water cup with a hand which is (soiled) with food,” thus the training is to be done.

@  eng-sk-56.tex
56. I shall not throw away bowl-washing water which has rice-grains (in it) in an inhabited area,” thus the training is to be done.

@ eng-sk-bhojana.tex
(Here ends) the Group of Thirty regarding Food.

@ eng-sk-57.tex
57. I shall not teach Dhamma to one who has a sunshade in (his) hand, (and) who is not ill,” thus the training is to be done.

@  eng-sk-58.tex
58. I shall not teach Dhamma to one who has a stick in (his) hand, (and) who is not ill,” thus the training is to be done.

@  eng-sk-59.tex
59. I shall not teach Dhamma to one who has a knife in (his) hand (and) who is not ill,” thus the training is to be done.

@  eng-sk-60.tex
60. I shall not teach Dhamma to one who has a weapon in (his) hand, (and) who is not ill,” thus the training is to be done.

@  eng-sk-61.tex
61. I shall not teach Dhamma to one who is wearing shoes, (and) who is not ill,” thus the training is to be done.

@  eng-sk-62.tex
62. I shall not teach Dhamma to one who is wearing sandals, (and) who is not ill,” thus the training is to be done.

@  eng-sk-63.tex
63. I shall not teach Dhamma to one who is in a vehicle, (and) who is not ill,” thus the training is to be done.

@  eng-sk-64.tex
64. I shall not teach Dhamma to one who is on a couch, (and) who is not ill,” thus the training is to be done.

@  eng-sk-65.tex
65. I shall not teach Dhamma to one sitting with (the knees) clasped-around, (and) who is not ill,” thus the training is to be done.

@  eng-sk-66.tex
66. I shall not teach Dhamma to one whose head is wrapped (with a turban), (and) who is not ill,” thus the
training is to be done.

@  eng-sk-67.tex
67. I shall not teach Dhamma to one whose head is covered, (and) who is not ill,” thus the training is to be done.

@  eng-sk-68.tex
68. Having sat down on the ground, I shall not teach Dhamma, to one who is sitting on a seat, (and) who is not ill,” thus the training is to be done.

@  eng-sk-69.tex
69. Having sat down on a low seat, I shall not teach Dhamma to one who is sitting on a high seat (and) who is not ill,” thus the training is to be done.

@  eng-sk-70.tex
70. I shall not teach Dhamma (while) standing, to one who is sitting, (and) who is not ill,” thus the training is to be done.

@  eng-sk-71.tex
71. I shall not teach Dhamma (while) walking behind, to one who is going in front, (and) who is not ill,” thus the training is to be done.

@  eng-sk-72.tex
72. I shall not teach Dhamma (while) walking off the path to one walking on the path, (and) who is not ill,” thus the training is to be done.

@ eng-sk-dhammadesana.tex
(Here ends) the Group of Sixteen regarding Teaching Dhamma.

@ eng-sk-73.tex
73. I shall not excrete or urinate (while) standing (and while) not ill,” thus the training is to be done.

@  eng-sk-74.tex
74. I shall not excrete or urinate or spit on crops, (while) not ill,” thus the training is to be done.

@  eng-sk-75.tex
75. I shall not excrete or urinate or spit in water, (while) not ill,” thus the training is to be done.

@ eng-sk-pakinnaka.tex
(Here ends) the Three Miscellaneous

@ eng-sk-uddittha.tex
Venerables, , the cases related to the training have been recited.

@ eng-sk-parisuddha.tex
Concerning that I ask the Venerables: (Are you) pure in this?\\
A second time again I ask: (Are you) pure in this?\\
A third time again I ask: (Are you) pure in this?\\
The Venerables are pure in this, therefore there is silence, thus I keep this (in mind).

@ eng-sk-nitthito.tex
The cases related to the training have finished.
