@ eng-pk-0.tex
Uposatha Preliminary Duties

@ eng-pk-nava.tex
Before doing the Uposatha, the ninefold preliminary duties should be done:

@ eng-pk-nava1.tex
sweeping the place;

@ eng-pk-nava2.tex
lighting the lanterns there;

@ eng-pk-nava3.tex
spreading the sitting mats;

@ eng-pk-nava4.tex
preparing water for drinking and using;

@ eng-pk-nava5.tex
bringing consent of the bhikkhus who are qualified to give their consent;

@ eng-pk-nava6.tex
bringing the purity of those bhikkhus who do not participate in the Uposatha;

@ eng-pk-nava7.tex
telling the season;

@ eng-pk-nava8.tex
counting the bhikkhus;

@ eng-pk-nava9.tex
instructing the bhikkhunis;

@ eng-pk-tattha.tex
Of these nine the first four have been done by the bhikkhus who know the bhikkhus' duties, these are therefore completed.

@ eng-pk-aharana.tex
But there is no bringing of consent and purity because there are in this boundary no bhikkhus who sit further from each other than the arm's length.

@ eng-pk-utukkhanam.tex
Telling the season is declaring the season thus: this number (of Uposathas) have passed, this number remain to come.

@ eng-pk-utunidha.tex
In this dispensation there are three seasons: winter, summer and rainy seasons.  This is the winter season and in this season there are: eigth Uposathas.  With this fortnight: one Uposatha has arrived, two Uposathas have past, five Uposathas are to come.  Thus the telling the season should be born in mind by you.

(Yes, venerable sir)

@ eng-pk-ganana.tex
The counting of bhikkhus is called the numbering of bhikkhus to such an amount who gather here in this Uposatha hall for the sake of the Uposatha. There are four bhikkhus gathered in this Uposatha hall.  Thus counting the bhikkhus should be born in mind by you.

(Yes, venerable sir)

@ eng-pk-ovado.tex
There is no instructing the bhikkhunis because of their non-existence now.

@ eng-pk-iti.tex
Thus the ninefold preliminary duties has been completed, since done is what there has been occasion to do and since that which there is no occasion to do has already been completed.

@ eng-pk-nitthite.tex
When the preliminary duty is completed:

@ eng-pk-nitthite0.tex
If the day is any one amongst the three, namely, the fourteenth, the fifteenth or the day of concordance; today being the fifteenth day Uposatha.

@ eng-pk-nitthite1.tex
1. Whatever number of bhikkhus there are who come to the function, who are appropriate for Community Uposatha, they may be four or more than that, if they are qualified, not having fallen into pārājika offences, not having been exluded by the Community, and

@ eng-pk-nitthite2.tex
2. they not having left the arms's length (not sitting further from each other than half an arm's distance), being in the same boundary, and

@ eng-pk-nitthite3.tex
3. if they have not committed the same offences such as taking food during the wrong hours, and

@ eng-pk-nitthite4.tex
4. if there is no disqualified person who should be avoided by making him sit beyond the arm's length (of the Community sitting) within the arm's length.

@ eng-pk-evantam.tex
Then the function of Uposatha endowed with these four characteristics is called 'reaching readiness' and is suitable to observe.

@ eng-pk-uposathakammassa.tex
The Community having known the 'reaching readiness' of the Uposatha-function they should understand that the Uposatha is being done now.

(Yes, venerable sir)

@ eng-pk-senior.tex
Senior bhikkhu: With the permission of the Community who has finished (all) preparations and preliminary functions and sat down here, may I invite (you) to recite the Pātimokkha.
