@ pali-8seasons.tex
...aṭṭha uposathā imina pakkhena eko uposatho sampatto... +

1st: satta uposathā avisiṭṭhā. +
2nd: eko uposatho atikkanto, cha uposathā avasiṭṭhā. +
3rd: dve uposathā atikkantā, pañca uposathā avasiṭṭhā. +
4th: tayo uposathā atikkantā, cattāro uposathā avasiṭṭhā. +
5th: cattāro uposathā atikkantā, tayo uposathā avasiṭṭhā. +
6th: pañca uposathā atikkantā, dve uposathā avasiṭṭhā. +
7th: cha uposathā atikkantā, eko uposatho avasiṭṭho. +
8th: satta uposathā atikkantā, aṭṭha uposathā paripuṇṇā. +

@ pali-7seasonswithPavarana.tex
...satta ca uposathā ekā ca pavāraṇā iminā pakkhena, eko uposatho sampatto... +

1st: cha ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
2nd: eko uposatho atikkanto, pañca ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
3rd: dve uposathā atikkantā, cattāro ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
4th: tayo uposathā atikkantā, tayo ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
5th: cattāro uposathā atikkantā, dve ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
6th: (see the separate section on the Pavāraṇā.) +
7th: pañca ca uposathā ekā ca pavāraṇā atikkantā, eko uposatho avasiṭṭho. +
8th: cha ca uposathā ekā ca pavāraṇā atikkantā, satta ca uposathā ekā ca pavāraṇā paripuṇṇā. +

@ pali-10seasons.tex
...adhikamāsavasena dasa uposathā, iminā pakkhena, eko uposatho sampatto... +

1st: nava uposathā avasiṭṭhā. +
2nd: eko uposatho atikkanto, aṭṭha uposathā avasiṭṭhā. +
3rd: dve uposathā atikkantā, satta uposathā avasiṭṭhā. +
4th: tayo uposathā atikkantā, cha uposathā avasiṭṭhā. +
5th: cattāro uposathā atikkantā, pañca uposathā avasiṭṭhā. +
6th: pañca uposathā atikkantā, cattāro uposathā avasiṭṭhā. +
7th: cha uposathā atikkantā, tayo uposathā avasiṭṭhā. +
8th: satta uposathā atikkantā, dve uposathā avasiṭṭhā. +
Ninth: aṭṭha uposathā atikkantā, eko uposatho avasiṭṭho. +
Tenth: nava uposathā atikkantā, dasa uposathā paripuṇṇā. +

@ pali-9seasonswithPavarana.tex
...adhikamāsavasena nava ca uposathā ekā ca pavāraṇā iminā pakkhena... +

1st: eko uposatho sampatto, aṭṭha ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
2nd: eko uposatho sampatto, eko uposatho atikkanto, satta ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
3rd: eko uposatho sampatto, dve uposathā atikkantā, cha ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
4th: eko uposatho sampatto, tayo uposathā atikkantā, pañca ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
5th:  eko uposatho sampatto, cattāro uposathā atikkantā, cattāro ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
6th:  eko uposatho sampatto, pañca uposathā atikkantā, tayo ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
7th: cha uposathā atikkantā, dve ca uposathā ekā ca pavāraṇā avasiṭṭhā. +
8th: (see the separate section on the Pavāraṇā.) +
9th: satta ca uposathā ekā ca pavāraṇā atikkantā, eko uposatho avasiṭṭho. +
10th: aṭṭha ca uposathā ekā ca pavāraṇā atikkantā, nava ca uposathā ekā ca pavāraṇā paripuṇṇā. +
